---
title: "EdgeCount Package Tutorial"  
author: "Joël R. Pradines"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
    number_sections: true
vignette: >
  %\VignetteIndexEntry{EdgeCount Package Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{css, echo=FALSE}
/* This rule highlights the reference links (e.g., [1]) */
a:target {
  background-color: #ffeb3b; /* A nice yellow color */
  transition: background-color 0.8s ease;
  border-radius: 3px;
  padding: 0 2px;
  outline: 1px solid #fbc02d;
}

/* This rule highlights only the HEADERS for section jumps */
div:target > h1,
div:target > h2,
div:target > h3,
div:target > h4,
div:target > h5,
div:target > h6 {
  background-color: #ffeb3b; /* A nice yellow color */
  transition: background-color 0.8s ease;
  display: inline-block; /* Ensures background only covers the text */
  padding: 2px 8px;
  border-radius: 4px;
}
```

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
library(EdgeCount)
library(data.table)
library(patchwork)

plot_graph <- function(g){
  ggraph(g) +
    geom_edge_link(color = "blue", width = 3) +
    geom_node_point(shape = 21, color = "black", size = 40, stroke = 1, fill="orange") +
    geom_node_text(aes(label = name), color = "black", size =20) +
    theme_void() +
    coord_cartesian(clip = "off") + 
    theme(plot.margin = unit(c(1, 1, 1, 1), "cm"))
}

plot_urchin_graph <- function(ecp, core_set, symbol_lookup = NULL) {
  
  if (!requireNamespace("igraph", quietly = TRUE) || !requireNamespace("ggraph", quietly = TRUE)) {
    message("Packages 'igraph' and 'ggraph' are required to plot the graph.")
    return(invisible(NULL))
  }
  
  # --- 1. Identify Nodes and Edges ---
  core_nodes <- intersect(core_set, ecp@names)
  if (length(core_nodes) == 0) return(invisible(NULL))

  adj_subset <- ecp@adj[core_nodes]
  
  internal_edges_list <- list()
  external_stubs_list <- list()
  stub_nodes <- c()

  for (node in core_nodes) {
    all_neighbors <- adj_subset[[node]]
    internal_neighbors <- intersect(all_neighbors, core_nodes)
    external_neighbors <- setdiff(all_neighbors, core_nodes)
    
    n_external <- length(external_neighbors)
    
    # Create internal edges (only one way to avoid duplication)
    if (length(internal_neighbors) > 0) {
      # Only add edges to nodes that are "after" the current one to avoid A-B and B-A
      internal_neighbors_filtered <- internal_neighbors[internal_neighbors > node]
      if(length(internal_neighbors_filtered) > 0) {
        internal_edges_list <- c(internal_edges_list, list(data.table(
          from = node,
          to = internal_neighbors_filtered
        )))
      }
    }
    
    # Create "stub" nodes and edges for external neighbors
    if (n_external > 0) {
      new_stub_nodes <- paste0(node, "_stub_", 1:n_external)
      stub_nodes <- c(stub_nodes, new_stub_nodes)
      
      external_stubs_list <- c(external_stubs_list, list(data.table(
        from = node,
        to = new_stub_nodes
      )))
    }
  }

  # --- 2. Create Node Metadata ---
  nodes_df <- data.frame(
    name = c(core_nodes, stub_nodes),
    category = c(rep("Core Set", length(core_nodes)), 
                 rep("Spike", length(stub_nodes))),
    stringsAsFactors = FALSE
  )
  
  nodes_df$label <- NA_character_ 
  
  if (!is.null(symbol_lookup)) {
    # Get the indices for the core nodes
    core_indices <- which(nodes_df$category == "Core Set")
    core_node_ids <- nodes_df$name[core_indices]
    
    symbols <- symbol_lookup[core_node_ids]
    # Fallback to the original ID if a symbol is not found (NA)
    nodes_df$label[core_indices] <- ifelse(is.na(symbols), core_node_ids, symbols)
  }

  # --- 3. Create Edge Metadata ---
  internal_edges <- rbindlist(internal_edges_list)
  external_edges <- rbindlist(external_stubs_list)
  
  if (nrow(internal_edges) > 0) internal_edges[, type := "Internal"]
  if (nrow(external_edges) > 0) external_edges[, type := "External"]
  
  all_edges_df <- rbind(internal_edges, external_edges)
  
  # --- 4. Create igraph Object ---
  g <- igraph::graph_from_data_frame(all_edges_df, directed = FALSE, vertices = nodes_df)
  
  ggraph(g) +
    # Draw edges
    geom_edge_link(
      aes(width = type, alpha = type, color = type)
    ) +
    scale_edge_width_manual(values = c("External" = 4, "Internal" = 6)) +
    scale_edge_alpha_manual(values = c("External" = 0.7, "Internal" = 1.0)) +
    scale_edge_color_manual(values = c("External" = "lightblue", "Internal" = "blue")) +
    
    
    # Draw nodes
    geom_node_point(
      aes(size = category, fill = category, color = category), 
      shape = 21,
      stroke = 4
    ) +
    scale_size_manual(values = c("Core Set" = 26, "Spike" = 0.01)) +
    scale_fill_manual(values = c("Core Set" = "orange", "Spike" = "white")) +
    scale_color_manual(values = c("Core Set" = "black", "Spike" = "white")) +
    
    geom_node_text(aes(label = label), # 'label' column now has NAs for spikes
                   repel = TRUE, 
                   color = "black",
                   bg.color = NA,
                   size = 15) +
    
    coord_cartesian(clip = "off") + 
    theme_void() +
    theme(
      legend.position = "none",
      plot.margin = unit(c(1, 1, 1, 1), "cm")
    )
}

plot_between_sets <- function(ecp, core_set_1, core_set_2, symbol_lookup = NULL) {

  if (!requireNamespace("igraph", quietly = TRUE) || !requireNamespace("ggraph", quietly = TRUE)) {
    message("Packages 'igraph' and 'ggraph' are required to plot the graph.")
    return(invisible(NULL))
  }

  # --- 1. Identify Node Sets ---
  nodes_set_1 <- intersect(core_set_1, ecp@names)
  nodes_set_2 <- intersect(core_set_2, ecp@names)

  inter <- intersect(nodes_set_1, nodes_set_2)
  nodes_set_1 <- setdiff(nodes_set_1, inter)
  nodes_set_2 <- setdiff(nodes_set_2, inter)

  all_core_nodes <- union(nodes_set_1, nodes_set_2)

  if (length(all_core_nodes) == 0) return(invisible(NULL))

  # --- 2. Initialize Edge & Node Lists ---
  intra_edges_list <- list()
  inter_edges_list <- list()
  external_stubs_list <- list()
  stub_nodes <- c()

  # --- 3. Loop over all core nodes to categorize edges ---
  for (node in all_core_nodes) {
    all_neighbors <- ecp@adj[[node]]

    # Get neighbors from all categories
    neighbors_in_set_1 <- intersect(all_neighbors, nodes_set_1)
    neighbors_in_set_2 <- intersect(all_neighbors, nodes_set_2)
    external_neighbors <- setdiff(all_neighbors, all_core_nodes)

    n_external <- length(external_neighbors)

    # --- Edge Logic ---
    # We only build edges *from* Set 1 to avoid duplication
    if (node %in% nodes_set_1) {
      # Add Intra-Set 1 edges
      if (length(neighbors_in_set_1) > 0) {
        # Filter for non-duplicates
        neighbors_in_set_1_filtered <- neighbors_in_set_1[neighbors_in_set_1 > node]
        if (length(neighbors_in_set_1_filtered) > 0) {
          intra_edges_list <- c(intra_edges_list, list(data.table(
            from = node, to = neighbors_in_set_1_filtered, type = "Intra-set"
          )))
        }
      }

      # Add Inter-Set edges (from Set 1 to Set 2)
      if (length(neighbors_in_set_2) > 0) {
        inter_edges_list <- c(inter_edges_list, list(data.table(
          from = node, to = neighbors_in_set_2, type = "Inter-set"
        )))
      }
    }

    # Add Intra-Set 2 edges (looping only from Set 2 nodes)
    if (node %in% nodes_set_2) {
      if (length(neighbors_in_set_2) > 0) {
        neighbors_in_set_2_filtered <- neighbors_in_set_2[neighbors_in_set_2 > node]
        if (length(neighbors_in_set_2_filtered) > 0) {
          intra_edges_list <- c(intra_edges_list, list(data.table(
            from = node, to = neighbors_in_set_2_filtered, type = "Intra-set"
          )))
        }
      }
    }

    # --- Stub Logic ---
    n_external <- 0
    if (n_external > 0) {
      new_stub_nodes <- paste0(node, "_stub_", 1:n_external)
      stub_nodes <- c(stub_nodes, new_stub_nodes)

      external_stubs_list <- c(external_stubs_list, list(data.table(
        from = node, to = new_stub_nodes, type = "External"
      )))
    }
  }

  # --- 4. Create Node Metadata ---
  nodes_df <- data.frame(
    name = c(nodes_set_1, nodes_set_2, stub_nodes),
    category = c(rep("Set 1", length(nodes_set_1)),
                 rep("Set 2", length(nodes_set_2)),
                 rep("Spike", length(stub_nodes))),
    stringsAsFactors = FALSE
  )

  nodes_df$label <- NA_character_
  if (!is.null(symbol_lookup)) {
    core_indices <- which(nodes_df$category %in% c("Set 1", "Set 2"))
    core_node_ids <- nodes_df$name[core_indices]

    symbols <- symbol_lookup[core_node_ids]
    nodes_df$label[core_indices] <- ifelse(is.na(symbols), core_node_ids, symbols)
  }

  # --- 5. Create Final Edge Table ---
  all_edges_df <- rbindlist(c(intra_edges_list, inter_edges_list, external_stubs_list))

  # --- 6. Create igraph Object ---
  g <- igraph::graph_from_data_frame(all_edges_df, directed = FALSE, vertices = nodes_df)


  ggraph(g) +
    # Draw edges
    geom_edge_link(
      aes(width = type, alpha = type, color = type)
    ) +
    # Define edge styles
    scale_edge_width_manual(values = c("Inter-set" = 2.0, "Intra-set" = 0.5, "External" = 0.5)) +
    scale_edge_alpha_manual(values = c("Inter-set" = 1.0, "Intra-set" = 0.8, "External" = 0.6)) +
    scale_edge_color_manual(values = c("Inter-set" = "red", "Intra-set" = "black", "External" = "gray")) +

    # Draw nodes
    geom_node_point(
      aes(size = category, fill = category, color = category),
      shape = 21,
      stroke = 1
    ) +
    scale_size_manual(values = c("Set 1" = 6, "Set 2" = 6, "Spike" = 0.01)) +
    scale_fill_manual(values = c("Set 1" = "#FFA500", "Set 2" = "#00BFFF", "Spike" = "white")) +
    scale_color_manual(values = c("Set 1" = "black", "Set 2" = "black", "Spike" = "white")) +

    # Add labels
    geom_node_text(aes(label = label),
                   repel = TRUE,
                   color = "black",
                   bg.color = NA,
                   bg.r = 0.15,
                   fontface = "bold",
                   size = 5) +

    coord_cartesian(clip = "off") +
    theme_void() +
    theme(
      legend.position = "none",
      plot.margin = unit(c(1, 1, 1, 1), "cm")
    )
}

```

```{css, echo=FALSE}
body {
  font-size: 100%;
  line-height: 1.6;
  max-width: 940px;
  margin: 0 auto;
  padding: 20px;
}
img {
  border: none;
}
```
# Introduction
The `EdgeCount` package provides a statistical framework for enrichment analysis based on graph connectedness. It is designed for large, sparse graphs, such as protein interaction networks, social networks, or term-element membership graphs (bipartite graphs).

The package's methodology combines the Random Graph with Given Expected Degrees (RGGED) model [[1]](#ref1) with analytical expressions for connectedness effect size and p-value [[2]](#ref2). 

It also introduces Vertex Set Enrichment Analysis (VSEA), a permutation-based method similar to GSEA [[3]](#ref3). VSEA uniquely accounts for the degree of each element (i.e., the number of terms an element belongs to).

The primary functions allow users to:

- Evaluate a graph's suitability for some of the EdgeCount statistical models.
- Test for internal connectedness within one or more sets.
- Test for connectedness between pairs of sets.
- Score a collection of terms against a single set or a data table of element sets.
- Perform Vertex Set Enrichment Analysis (VSEA) on a ranked list of elements.
- Summarize VSEA results with Normalized Enrichment Scores (NES) and FDR q-values.

<br>

# Installation

The `EdgeCount` package is currently available on GitHub. You can install it using the `devtools` package. If you do not have `devtools` installed, you can install it from CRAN.

```{r installation, eval=FALSE}
# install.packages("devtools")
devtools::install_github("JoelRMath/EdgeCount")
```

<br>

# Usage

This section provides an overview of the main use cases for the package.

## Creating a Graph

The core objects for any analysis are `ECGraph`, which holds the graph structure, and `ECProb`, which adds statistical model parameters and methods.

The first step is to create these objects from an edge list. The `ECGraph()` constructor is flexible and can accept this list as a `data.frame`, a file path, or an `igraph` object.

Here is an example for a graph of order 5 (number of vertices) and size 5 (number of edges).


```{r create_ECGraph_ECProb, message=FALSE, warning=FALSE}
library(EdgeCount)
library(data.table)
library(igraph)

# Data frame of edges 
edge_df <- data.frame(
  p1 = c("A", "B", "C", "D", "E"),
  p2 = c("B", "C", "A", "E", "A"),
  stringsAsFactors = FALSE
)

# ECGraph and ECProb objects from the edge list
ecg <- ECGraph(edge_df)
ecp <- ECProb(ecg)

# Same from a file
temp_file <- tempfile(fileext = ".tsv")
write.table(edge_df, file = temp_file, sep = "\t", row.names = FALSE, quote = FALSE)
ecg_from_file <- ECGraph(temp_file, col1 = "p1", col2 = "p2", header = TRUE, sep = "\t")
ecp_from_file <- ECProb(ecg_from_file)
unlink(temp_file)

# Same from an igraph object
igraph_obj <- igraph::graph_from_data_frame(edge_df, directed = FALSE)
ecg_from_igraph <- ECGraph(igraph_obj)
ecp_from_igraph <- ECProb(ecg_from_igraph)

# ECProb structure
str(ecp)
```

The S4 class `ECProb` inherits the slots of `ECGraph`:

 - `@names`: vertices names, as a chr vector
 - `@adj`: adjacency list, as a named list linking each vertex to its neighbors
 - `@degrees`: vertex degrees, as a named list
 
`ECProb` has the additional slots `@graph_size`, `@graph_order` and `@average_degree`, as well as statistical methods and functions that will be presented later. In order to properly use these functions, it is important to first check the suitability of the underlying `ECGraph` for some of them.

## Suitability for fast approximations

Even though it is not based on a large `ECGraph`, the previous example of `ECProb` object meets the criteria for fast approximations.

```{r plot_good_graph, message=FALSE, warning=FALSE, echo=FALSE, out.width="20%", fig.width=8, fig.height = 8, fig.align = "center"}

library(igraph)
library(ggraph)

# 1. Create the igraph object from the edge list
edge_df <- data.frame(
  p1 = c("A", "B", "C", "D", "E"),
  p2 = c("B", "C", "A", "E", "A"),
  stringsAsFactors = FALSE
)
g <- graph_from_data_frame(edge_df, directed = FALSE)

# 2. Create the plot
set.seed(1)
plot_graph(g)
```

This is verified by using method `summarize_suitability_fast` of `ECProb`.

```{r suitability_sparse}
suitability_stats <- summarize_suitability_fast(ecp)
suitability_stats
```

- `$pij_over_1` is the important criterion: the proportion of edges (among all possible pairs $(i,j)$ of vertices) that yield a connection probability $p_{ij}$ greater than 1 (hence unsuitable) under an approximate analytical framework of the RGGED. A value close to 0 means that fast approximation methods are appropriate.
- `$prop_problematic_vertices` is the proportion of vertices with degree greater than $\sqrt{2M}$, where $M$ is the graph size `ecp@graph_size`. Indeed, under the model for fast approximations, $p_{ij} = k_ik_j/(2M)$, where $k$ is a vertex degree; so any vertex with $k_i > \sqrt{2M}$ might yield  $p_{ij} > 1$ for connection to another problematic vertex $j$.
- `$summary_pij` is a short summary of the distribution of connection probabilities over all pairs of vertices
- `$pij_distribution` is the full distribution of $p_{ij}$ values in the RGGED model with fast approximations.

Here is now an example of small graph that, even though not a complete graph, does not pass the suitability test.

```{r plot_bad_graph, message=FALSE, warning=FALSE, echo=FALSE, out.width="20%", fig.width=8, fig.height = 8, fig.align = "center"}

edge_df <- data.frame(
  p1 = c("A", "A", "A", "A", "A", "B", "B", "B", "B", "C", "C", "C"),
  p2 = c("B", "C", "D", "E", "F", "C", "G", "H", "I", "J", "K", "L"),
  stringsAsFactors = FALSE
  )
g <- graph_from_data_frame(edge_df, directed = FALSE)

set.seed(2)
plot_graph(g)
```

```{r suitability_non_sparse}
edge_df <- data.frame(
  p1 = c("A", "A", "A", "A", "A", "B", "B", "B", "B", "C", "C", "C"),
  p2 = c("B", "C", "D", "E", "F", "C", "G", "H", "I", "J", "K", "L"),
  stringsAsFactors = FALSE
)
ecp <- ECProb(ECGraph(edge_df))
suitability_stats <- summarize_suitability_fast(ecp)
suitability_stats
```


The `pij_over_1` value is 4.5%. This indicates that some vertex pairs have an approximate edge probability greater than or equal to 1, making the fast approximation methods unsuitable. For such a graph, other methods are available in the package.

In all rigor the fast approximation methods based on $p_{ij} = k_ik_j/(2M)$ are valid when $p_{ij} \ll 1$, otherwise $p_{ij} = 1-\exp(-k_ik_j/(2M))$ or $p_{ij} = \min(1, k_ik_j/(2M))$ need to be utilized, and the resulting methods are slower. 

**However, the fast approximation methods are conservative**:

- They overestimate $p_{ij}$ values.
- They underestimate effect sizes.
- They overestimate edge-count p-values.

**That is, fast approximation methods cannnot yield false-positive results**. In other words, using the fast approximation methods by default is a conservative approach for analysis.

The package comes with object `sample_ecg`, an example of real large sparse graph based on the human network of the STRING database (version 12.0)[[4]](#ref4),[[5]](#ref5). The network is limited to protein interactions having a confidence score greater than or equal to 900. Proteins IDs have been replaced with their mapped Gene Entrez IDs. 


```{r sample_ecg}
data("sample_ecg")
data("sample_ects")
sample_ecp <- ECProb(sample_ecg)
str(sample_ecp, list.len = 6)
```

This network is relatively large and it is sparse.

```{r sample_ecg_suitability}
summarize_suitability_fast(sample_ecp)
```

Indeed, it is suitable for fast approximations not only because `pij_over_1` = 0, but also because the majority of $p_{ij}$ values are close to $4.10^{-4}$. 

Object `sample_ecp` is utilized next to provide examples of `ECProb` uses.

## Analysis of Interaction Networks (ECProb)

Example of connectedness analyses within an undirected, unipartite graph are now presented.

### Tests for *within-set* connectedness

This test answers the question: "Are the vertices of a given set $S$ more connected to each other than expected by *chance*?" The concept of *chance* is here based on the RGGED: the distribution of the number of edges within set $S$ over all possible realizations of the RGGED, which in average preserve degrees of all vertices. Using such a null model that accounts for individual vertex degrees is statistically important.

#### Basic method 

The package comes with a sample bipartite network `sample_ects` that maps a trimmed version of GO categories to their human gene elements [[6]](#ref6), [[7]](#ref7), [[9]](#ref9). Object `sample_ects@ecprob` is utilized below to define with a term a set of elements/vertices which is then tested for its connectedness in interaction network `sample_ecp`.

```{r setA_in, message=FALSE, warning=FALSE}
# bipartite network (term-element relationships)
data("sample_ects")
# term descriptions
data("sample_term_lookup")
setA_id <- "GO:0032395"
setA_name <- sample_term_lookup[setA_id]
setA_name
# elements mapped to term (bipartite network sample_ects@ecprob)
setA <- get_neighbors(sample_ects@ecprob, setA_id) 
length(setA)
# observed edge count in the graph between elements (interaction network sample_ecg)
data("sample_ecg")
sample_ecp <- ECProb(sample_ecg)
observed_ec <- get_edge_count_in(sample_ecp, setA) 
# calculation of statistics
statsA <- edge_count_statistics(sample_ecp, 
                               set1 = setA, 
                               set2 = NULL, 
                               observed_edge_count = observed_ec,
                               lambda_method = "fast")
statsA
```

Method `edge_count_statistics` returns a list:

- `$observed_edge_count` $x$ is the observed number of edges, here the edge count within `setA`
- `max_possible_edges` $m$ is the maximum number of edges one could observe in the RGGED, here the number of pairs withing `setA`, i.e. $(7\times6)/2 = 21$
- `$lambda`, $\lambda$ is the expected number of observed edges within `setA` in the RGGED, that is, the average number of edges over all realizations of the RGGED
- `$p-value` $p_S$ is $\Pr\left( X \geq x\right)$, where $X$ is the random variable 'number of edges within `setA`' in the RGGED
- `$log2_Anscombe_ratio` $lAr$ is the observed *log effect size*
- `$size1` is the size of the input set after it has been reduced to vertices represented in the `ECProb` object
- `$size2` is used only for between-sets connectedness. Here it is zero because argument `set2` is null

Calling $p_{ij}$ the probability of connecting two vertices $i$ and $j$ of a set $S$ in the RGGED, $k$ the degree of a vertex, and $M$ the graph size we have in general

$$p_{ij} = \min\left(1, \frac{k_ik_j}{2M}\right)$$
$$m = \frac{1}{2}\left| S \right|\left( \left| S\right|-1\right)$$
$$\lambda = \sum_{i < j, \in S} p_{ij}$$
$$p_S = \sum_{u=x}^m \Pr\left(X = u\right)$$
$$\Pr\left(X = u\right) = \frac{1}{\alpha}\frac{\lambda^u}{u!}$$
$$\alpha = \sum_{v=0}^{m} \frac{\lambda^v}{v!}$$
$$lAr = \frac{1}{2} \log_2 \frac{x +3/8}{\lambda + 3/8}$$
Using $lAr$ as effect size is important because $X$ can have a truncated Poisson distribution of parameter $\lambda$ small enough to yield an asymmetric distribution.

With `lambda_method = "fast"` calculations are simplified

$$p_{ij} = \frac{k_ik_j}{2M} \quad \text{and} \quad \lambda = \frac{1}{4M} \left( \left(\sum_{i\in S} k_i\right)^2 - \sum_{i\in S} k_i^2\right)$$
Namely, time complexity is now $O\left( \left| S \right|\right)$ instead of $O\left( \left| S \right|^2\right)$.

To illustrate the importance of accounting for the degree of each vertex consider now the following `setB`.

```{r setB_in, message=FALSE, warning=FALSE}
# bipartite network (term-element relationships)
data("sample_ects")
setB_id <- "GO:0006729"
# term descriptions
data("sample_term_lookup")
setB_name <- sample_term_lookup[setB_id]
setB_name
# elements mapped to term (bipartite network sample_ects@ecprob)
setB <- get_neighbors(sample_ects@ecprob, setB_id) 
length(setB)
# observed edge count in the graph between elements (interaction network sample_ecg)
data("sample_ecg")
sample_ecp <- ECProb(sample_ecg)
observed_ec <- get_edge_count_in(sample_ecp, setB) 
# calculation of statistics
statsB <- edge_count_statistics(sample_ecp, 
                               set1 = setB, 
                               set2 = NULL, 
                               observed_edge_count = observed_ec,
                               lambda_method = "fast")
statsB
```

While `setA` and `setB` have the same size, and thus the same `max_possible_edges`, and the same `observed_edge-count`, `setB` has lesser connectedness than `setA`, as attested by a larger `p-value` and a smaller `log2_Anscombe_ratio`. This is due to `setB` having a larger `lambda` than `setA` because its vertices have larger degrees. This is illustrated below by plotting the subgraphs induced by the two sets while complementing visualization of their degrees in the entire network in the form of stubs.


```{r plot_urchin_comparison, message=FALSE, warning=FALSE, echo=FALSE, fig.width=25, fig.height=12, out.width="80%", fig.align = "center"}
library(patchwork)

tsz <- 35

# 1. Create Plot A and save it to a variable
p_A <- plot_urchin_graph(sample_ecp, setA, sample_gene_symbol_lookup) + 
  ggtitle(paste0("Set A: ", setA_id, "\n", setA_name, "\n",
                 "set_size = ", length(setA), 
                 ", observed_ec = ", statsA$observed_edge_count, "\n",
                 "p_value = ", signif(statsA$p_value, digits = 3), "\n",
                 "log2_Anscombe_ratio = ", signif(statsA$log2_Anscombe_ratio, digits = 3))) + 
  theme(plot.title = element_text(size = tsz))

# 2. Create Plot B and save it to a variable
p_B <- plot_urchin_graph(sample_ecp, setB, sample_gene_symbol_lookup) + 
  ggtitle(paste0("Set B: ", setB_id, "\n", setB_name, "\n",
                 "set_size = ", length(setB), 
                 ", observed_ec = ", statsB$observed_edge_count, "\n",
                 "p_value = ", signif(statsB$p_value, digits = 3), "\n",
                 "log2_Anscombe_ratio = ", signif(statsB$log2_Anscombe_ratio, digits = 3)))  + 
  theme(plot.title = element_text(size = tsz))

# 3
p_A + p_B
```

#### Vectorized method

The last examples illustrated how to test an individual set for its within connectedness. When dealing with multiple sets, it is better to utilize a vectorized method which has been optimized for speed and takes a data table of sets as argument. **Note that this method is specific to the fast lambda approximation**.

Method `calculate_in_stats_fast_vectorized(ecprob, sets_dt, set_membership_dt)` has the following arguments

- `ecprob`: an instance of `ECProb` (interaction network)
- `sets_dt`: a data table with a single column ("set_id") containing IDs of sets to test
- `set_membership_dt': a data table with "set_id" and "element" columns (bipartite network)

Below is the application of this method to the entire collection of sets  contained in `sample_ects` (GO categories [[6]](#ref6),[[7]](#ref7), [[8]](#ref8), [[9]](#ref9)) and the interaction network `sample_ecg` (extracted from STRING [[4]](#ref4),[[5]](#ref5)).


```{r in_vectorized}
# interaction network and associated ECProb
data("sample_ecg")
ecprob <- ECProb(sample_ecg)
# bipartite graph: term-element relationships
data("sample_ects")
set_membership_dt <- as.data.table(to_dataframe(sample_ects))
setnames(set_membership_dt, c("term", "element"), c("set_id", "element"))
sets_to_test <- data.table(set_id = unique(set_membership_dt$set_id))
length(sets_to_test$set_id)
length(set_membership_dt$element)
# within conncectedness statistics for all sets
system.time(results <- calculate_in_stats_fast_vectorized(ecprob, sets_to_test, set_membership_dt))
# adding annotation (term names, not just ids)
data("sample_term_lookup")
lookup_dt <- data.table(
  set_id = names(sample_term_lookup),
  term_name = sample_term_lookup
)
setkey(results, set_id)
setkey(lookup_dt, set_id)
annotated_results <- lookup_dt[results, on = "set_id"]
setorder(annotated_results, -log2_Anscombe_ratio)

print(annotated_results)

```

Note that `set_size` corresponds to that of the input set reduced to vertices present in the `ECProb` object. Method `calculate_in_stats_fast_vectorized` has been optimized for speed and its returned value is a data table containing edge-count statistics. Distributions of some of those statistics obtained with the above example are shown below.

```{r plot_distributions, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=8, fig.align="center", out.width="90%"}
# Load plotting libraries
library(ggplot2)
library(patchwork)

# 1. Plot for Observed Edge Count (log10 scale)
# We add +1 for the log-transform to handle the many 0s.
p1 <- ggplot(annotated_results, aes(x = log10(observed_edge_count + 1))) +
  geom_histogram(bins = 30, fill = "#0c4a6e", color = "white", alpha = 0.8) +
  ggtitle("Observed Edge Count") +
  xlab("log10(observed_edge_count + 1)") +
  theme_minimal(base_size = 14)

# 2. Plot for Lambda (Expected Count) (log10 scale)
# We filter for lambda > 0 for the log-transform.
p2 <- ggplot(annotated_results[lambda >0 ], aes(x = log10(lambda))) +
  geom_histogram(bins = 30, fill = "#047857", color = "white", alpha = 0.8) +
  ggtitle("Expected Edge Count") +
  xlab("log10(lambda > 0)") +
  theme_minimal(base_size = 14)

# 3. Plot for P-Values
# This histogram is key. A good result will have a spike at 0 (significant terms).
p3 <- ggplot(annotated_results, aes(x = p_value)) +
  geom_histogram(bins = 20, boundary = 0, fill = "#c2410c", color = "white", alpha = 0.8) +
  ggtitle("P-Value") +
  xlab("p_value") +
  theme_minimal(base_size = 14)

# 4. Plot for Log2 Anscombe Ratio (Effect Size)
p4 <- ggplot(annotated_results, aes(x = log2_Anscombe_ratio)) +
  geom_histogram(bins = 40, fill = "#9a3412", color = "white", alpha = 0.8) +
  ggtitle("Effect Size") +
  xlab("log2_Anscombe_ratio") +
  theme_minimal(base_size = 14)

# Combine all four plots into a 2x2 grid
(p1 + p2) / (p3 + p4)
```

### Test for *between-set* connectedness

This test answers the question: "Are two given vertex sets $S_1$ and $S_2$ more connected to each other than expected by *chance*?" The concept of *chance* is again based on the RGGED: the distribution of the number of edges between $S_1$ and $S_2$ over all possible realizations of the RGGED, a null model accounting for individual vertex degrees. This 'between-sets' connectedness is different from the 'within-connectedness' of individual sets. Therefore, the methods first reduce pairs of sets to their disjoint parts.

#### Basic method

The basic method is again `edge_count_statistics`. Its argument `observed_edge_count` is now obtained with method `get_edge_count_between`, which takes the two sets $S_1$ and $S_2$ as arguments.  

```{r ecprob_between_example}
line <- 316
df <- read.table("../data-raw/res/between_top10000.tsv", sep="\t", header = TRUE, quote = "", stringsAsFactors = FALSE)
setA_id <- df[line, 1]
setB_id <- df[line, 3]
# bipartite network (term-element relationships)
data("sample_ects")
data("sample_term_lookup")
setA_name <- sample_term_lookup[setA_id]
print(setA_name)
setB_name <- sample_term_lookup[setB_id]
print(setB_name)
setA <- get_neighbors(sample_ects@ecprob, setA_id) 
setB <- get_neighbors(sample_ects@ecprob, setB_id) 
# observed edge count in the graph between elements (interaction network sample_ecg)
data("sample_ecg")
sample_ecp <- ECProb(sample_ecg)
plot_between_sets(sample_ecp, setA, setB)

```

## Analysis of Bipartite Graphs (ECTermScoring)

This workflow is for analyzing term-element membership graphs (e.g., GO terms to genes). The main object for this is `ECTermScoring`. This is the foundation for Vertex Set Enrichment Analysis (VSEA).

```{r ects_example, eval=FALSE}
# This example is commented out to speed up the vignette build.
# You can run it manually in your console.

# 1. Load the sample ECTermScoring object
data("sample_ects")

# 2. Create a ranked list of elements
#    (Here we use a random ranking for demonstration)
set.seed(123)
ranked_elements <- sample(sample_ects@elements)
element_ranks <- setNames(seq_along(ranked_elements), ranked_elements)

# 3. Run the full VSEA analysis
#    (We use a small number of permutations for a quick example)
vsea_results <- run_vsea_analysis(
  sample_ects,
  element_ranks,
  n_permutations = 100,
  seed = 123
)

# 4. View the top results for positive enrichment
print(head(vsea_results$max_score_summary))
```

<br>

# References

<a id="ref1">(1.)</a> Chung & Lu (2002). The average distances in random graphs with given expected degrees. *PNAS*, 99 (25) 15879-1582. 
<a href="https://www.pnas.org/doi/10.1073/pnas.252631999" target="_blank">[link]</a>

<a id="ref2">(2.)</a> Pradines, J.R., et al. (2005). Analyzing Protein Lists with Large Networks: Edge-Count Probabilities in Random Graphs with Given Expected Degrees. *Journal of Computational Biology*, 12(2), 113-28.<a href="https://www.liebertpub.com/doi/10.1089/cmb.2005.12.113" target="_blank">[link]</a>


<a id="ref3">(3.)</a> Subramanian, A., et al. (2005). Gene set enrichment analysis: a knowledge-based approach for interpreting genome-wide expression profiles. *PNAS*, 102(43), 15545-50. <a href="https://www.pnas.org/doi/10.1073/pnas.0506580102" target="_blank">[link]</a>

<a id="ref4">(4.)</a> Szklarczyk D, Kirsch R, Koutrouli M, Nastou K, Mehryary F, Hachilif R, Annika GL, Fang T, Doncheva NT, Pyysalo S, Bork P, Jensen LJ, von Mering C. The STRING database in 2023: protein–protein association networks and functional enrichment analyses for any sequenced genome of interest. Nucleic Acids Res. 2023 Jan 6;51(D1):D638-646 <a href="https://pubmed.ncbi.nlm.nih.gov/36370105/" target= "_blank">[link]</a>

<a id="ref5">(5.)</a> Object `sample_ecg` is a subset of the human STRING network (version 12.0) with interaction scores of 900 or higher. Vertex IDs are Entrez Gene IDs [[8]](#ref8). The network has been trimmed to create a smaller, more manageable example for the package's vignettes and tests. Source: <a href="https://string-db.org/" target= "_blank">[string-db.org]</a>. Data from STRING is licensed under <a href="https://creativecommons.org/licenses/by/4.0/" target= "_blank">[CC BY 4.0]</a>.

<a id="ref6">(6.)</a> Ashburner et al. (2000). Gene ontology: tool for the unification of biology. *Nat Genet*, 25(1), 25-9. <a href="https://www.nature.com/articles/ng0500_25" target="_blank">[link]</a>

<a id="ref7">(7.)</a> The Gene Ontology Consortium. (2023). The Gene Ontology knowledgebase in 2023. *Genetics*, 224(1). <a href="https://doi.org/10.1093/genetics/iyad031" target="_blank">[link]</a>

<a id="ref8">(8.)</a> Carlson, M. (2019). org.Hs.eg.db: Genome wide annotation for Human. *R package*.

<a id="ref10">(9.)</a> `sample_ects` is derived from Gene Ontology (GO) term-to-gene mappings [[6, 7, 9]](#ref6). Element IDs are Entrez Gene IDs [[8]](#ref8). Source: <a href="http://geneontology.org/" target= "_blank">[geneontology.org]</a>. GO data is licensed under <a href="https://creativecommons.org/licenses/by/4.0/" target= "_blank">[CC BY 4.0]</a>.
