---
title: "EdgeCount Package Tutorial"  
author: "Joël R. Pradines"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{EdgeCount Package Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{css, echo=FALSE}
/* This rule highlights the reference links (e.g., [1]) */
a:target {
  background-color: #ffeb3b; /* A nice yellow color */
  transition: background-color 0.8s ease;
  border-radius: 3px;
  padding: 0 2px;
  outline: 1px solid #fbc02d;
}

/* This rule highlights only the HEADERS for section jumps */
div:target > h1,
div:target > h2,
div:target > h3,
div:target > h4,
div:target > h5,
div:target > h6 {
  background-color: #ffeb3b; /* A nice yellow color */
  transition: background-color 0.8s ease;
  display: inline-block; /* Ensures background only covers the text */
  padding: 2px 8px;
  border-radius: 4px;
}
```

```{r setup, include=FALSE}
library(EdgeCount)
library(data.table)
```

```{css, echo=FALSE}
body {
  font-size: 115%;
  line-height: 1.6;
  max-width: 940px;
  margin: 0 auto;
  padding: 20px;
}
```
# Introduction
The EdgeCount package provides a statistical framework for enrichment analysis based on graph connectedness. It is designed for large, sparse graphs, such as protein interaction networks, social networks, or term-element membership graphs (bipartite graphs).

The package's methodology combines the Random Graph with Given Expected Degrees (RGGED) model [[1]](#ref1) with analytical expressions for connectedness effect size and p-value [[2]](#ref2). 

It also introduces Vertex Set Enrichment Analysis (VSEA), a permutation-based method similar to GSEA [[3]](#ref3). VSEA uniquely accounts for the degree of each element (i.e., the number of terms an element belongs to).

The primary functions allow users to:

- Evaluate a graph's suitability (e.g., sparseness) for the EdgeCount statistical model.
- Test for internal connectedness within one or more sets.
- Test for connectedness between pairs of sets.
- Score a collection of terms against a single set or a vector of element sets.
- Perform Vertex Set Enrichment Analysis (VSEA) on a ranked list of elements.
- Summarize VSEA results with Normalized Enrichment Scores (NES) and FDR q-values.

<br>

# Installation

The EdgeCount package is currently available on GitHub. You can install it using the devtools package. If you do not have devtools installed, you can install it from CRAN.

```{r installation, eval=FALSE}
# install.packages("devtools")
devtools::install_github("JoelRMath/EdgeCount")
```

<br>

# Basic Usage

This section provides a quick overview of the main use cases for the `EdgeCount` package.

## Creating a Graph and Checking Suitability

The core objects for any analysis are `ECGraph` (which holds the graph structure) and `ECProb` (which adds the statistical model parameters).

The first step in any analysis is to create these objects from an edge list and check if the graph is suitable for the "fast" lambda approximation, which is used in many vectorized functions.

```{r}
library(EdgeCount)
library(data.table)

# 1. Start with a data frame of edges (e.g., a sparse graph)
edge_df <- data.frame(
  p1 = c("A", "B", "C", "D", "E"),
  p2 = c("B", "C", "A", "E", "A"),
  stringsAsFactors = FALSE
)

# 2. Create the ECGraph and ECProb objects
ecg <- ECGraph(edge_df)
ecp <- ECProb(ecg)

# 3. Test the graph's suitability for the "fast" lambda approximation
#    We use the summarize_suitability_fast() method.
suitability_stats <- summarize_suitability_fast(ecp)
print(suitability_stats)
```

The most important value here is `pij_over_1`. This estimates the proportion of vertex pairs for which the "fast" approximation (which assumes `p_ij < 1`) might be inaccurate. A value very close to 0, as seen here, indicates the graph is sparse and the "fast" lambda methods are suitable.

Conversely, let's look at a "dense" graph, such as a small complete graph where every node is connected to every other node.

```{r }
# 1. Create a not-so-sparse graph

edge_df <- data.frame(
  p1 = c("A", "A", "A", "A", "A", "B", "B", "B", "B", "C", "C", "C"),
  p2 = c("B", "C", "D", "E", "F", "C", "G", "H", "I", "J", "K", "L"),
  stringsAsFactors = FALSE
)

# 2. Create the ECProb object
ecp <- ECProb(ECGraph(edge_df))

# 3. Test its suitability
suitability_stats <- summarize_suitability_fast(ecp)
print(suitability_stats)
```

Here, the `pij_over_1` value is `1` (or 100%). This indicates that all vertex pairs have an expected edge probability greater than or equal to 1, making the "fast" lambda approximation completely unsuitable. For such graphs, the "accurate" or "optimized" lambda methods must be used.

## Analysis of Interaction Networks (ECProb)

This workflow is for analyzing connectedness within a standard, unipartite graph (like the `ecp` object we just created).

### Test for *within-set* connectedness

This analysis answers the question: "Are the elements within my set more connected to each other than expected by chance?"

```{r ecprob_within_example}
# We will re-use the sparse 'ecp' object from the previous step.

# 1. Define your set of interest
my_set <- c("A", "B", "C")

# 2. Calculate statistics
#    - First, get the observed edge count
obs_ec <- get_edge_count_in(ecp, my_set)
print(paste("Observed edges:", obs_ec))

#    - Then, get the full statistics
stats_in <- edge_count_statistics(
  ecp,
  set1 = my_set,
  observed_edge_count = obs_ec,
  lambda_method = "fast"
)
print(stats_in)
```

### Test for *between-set* connectedness

This analysis answers the question: "Are my two sets more connected to each other than expected by chance?"

```{r ecprob_between_example}
# (Continuing from the example above)

# 1. Define two sets of interest
set_1 <- c("A", "B")
set_2 <- c("D", "E")

# 2. Calculate statistics
#    - First, get the observed edge count
obs_ec <- get_edge_count_between(ecp, set_1, set_2)
print(paste("Observed edges:", obs_ec))

#    - Then, get the full statistics
stats_between <- edge_count_statistics(
  ecp,
  set1 = set_1,
  set2 = set_2,
  observed_edge_count = obs_ec,
  lambda_method = "fast"
)
print(stats_between)
```

## Analysis of Bipartite Graphs (ECTermScoring)

This workflow is for analyzing term-element membership graphs (e.g., GO terms to genes). The main object for this is `ECTermScoring`. This is the foundation for Vertex Set Enrichment Analysis (VSEA).

```{r ects_example, eval=FALSE}
# This example is commented out to speed up the vignette build.
# You can run it manually in your console.

# 1. Load the sample ECTermScoring object
data("sample_ects")

# 2. Create a ranked list of elements
#    (Here we use a random ranking for demonstration)
set.seed(123)
ranked_elements <- sample(sample_ects@elements)
element_ranks <- setNames(seq_along(ranked_elements), ranked_elements)

# 3. Run the full VSEA analysis
#    (We use a small number of permutations for a quick example)
vsea_results <- run_vsea_analysis(
  sample_ects,
  element_ranks,
  n_permutations = 100,
  seed = 123
)

# 4. View the top results for positive enrichment
print(head(vsea_results$max_score_summary))
```

<br>

# References

<a id="ref1">(1.)</a> Chung & Lu (2002). The average distances in random graphs with given expected degrees. *PNAS*, 99 (25) 15879-1582. 
<a href="https://www.pnas.org/doi/10.1073/pnas.252631999" target="_blank">[link]</a>

<a id="ref2">(2.)</a> Pradines, J.R., et al. (2005). Analyzing Protein Lists with Large Networks: Edge-Count Probabilities in Random Graphs with Given Expected Degrees. *Journal of Computational Biology*, 12(2), 113-28.<a href="https://www.liebertpub.com/doi/10.1089/cmb.2005.12.113" target="_blank">[link]</a>


<a id="ref3">(3.)</a> Subramanian, A., et al. (2005). Gene set enrichment analysis: a knowledge-based approach for interpreting genome-wide expression profiles. *PNAS*, 102(43), 15545-50. <a href="https://www.pnas.org/doi/10.1073/pnas.0506580102" target="_blank">[link]</a>

