---
title: "EdgeCount Package Tutorial"  
author: "Joël R. Pradines"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{EdgeCount Package Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{css, echo=FALSE}
/* This rule highlights the reference links (e.g., [1]) */
a:target {
  background-color: #ffeb3b; /* A nice yellow color */
  transition: background-color 0.8s ease;
  border-radius: 3px;
  padding: 0 2px;
  outline: 1px solid #fbc02d;
}

/* This rule highlights only the HEADERS for section jumps */
div:target > h1,
div:target > h2,
div:target > h3,
div:target > h4,
div:target > h5,
div:target > h6 {
  background-color: #ffeb3b; /* A nice yellow color */
  transition: background-color 0.8s ease;
  display: inline-block; /* Ensures background only covers the text */
  padding: 2px 8px;
  border-radius: 4px;
}
```

```{r setup, include=FALSE}
library(EdgeCount)
library(data.table)

plot_set_neighbors <- function(ecp, vertex_set) {
  
  if (!requireNamespace("igraph", quietly = TRUE) || !requireNamespace("ggraph", quietly = TRUE)) {
    message("Packages 'igraph' and 'ggraph' are required to plot the graph.")
    return(invisible(NULL))
  }
  
  # 1. Identify all nodes involved
  set_nodes <- intersect(vertex_set, ecp@names)
  neighbor_nodes <- get_neighbors(ecp, set_nodes)
  neighbor_nodes_outside <- setdiff(neighbor_nodes, set_nodes)
  all_nodes_in_plot <- c(set_nodes, neighbor_nodes_outside)
  
  if (length(all_nodes_in_plot) == 0) return(invisible(NULL))

  # 2. Create the node metadata (Categories 1 and 2)
  nodes_df <- data.frame(
    name = all_nodes_in_plot,
    stringsAsFactors = FALSE
  )
  nodes_df$category <- ifelse(nodes_df$name %in% set_nodes, "Set Member", "Neighbor")
  
  # 3. Get all edges for this subgraph
  adj_subset <- ecp@adj[set_nodes]
  edges_df <- data.table::rbindlist(lapply(adj_subset, function(n) data.table(to = n)), idcol = "from")
  
  # Filter edges to only those connecting nodes in our plot
  edges_df <- edges_df[to %in% all_nodes_in_plot]
  
  # 4. Create the igraph object
  g <- igraph::graph_from_data_frame(edges_df, directed = FALSE, vertices = nodes_df)
  
  # 5. Set edge types for styling
  # Get the categories of both ends of each edge
  ends_matrix <- igraph::ends(g, igraph::E(g))
  category1 <- igraph::V(g)$category[match(ends_matrix[,1], igraph::V(g)$name)]
  category2 <- igraph::V(g)$category[match(ends_matrix[,2], igraph::V(g)$name)]
  
  # Assign type
  edge_type <- ifelse(category1 == "Set Member" & category2 == "Set Member", "Internal", "External")
  igraph::E(g)$type <- edge_type

  # 6. Plot with ggraph
  set.seed(2) # for reproducible layout
  ggraph::ggraph(g) +
    # Draw edges
    ggraph::geom_edge_link(
      aes(width = type, alpha = type, color = type)
    ) +
    # Define edge styles
    ggraph::scale_edge_width_manual(values = c("External" = 0.5, "Internal" = 1.5)) +
    ggraph::scale_edge_alpha_manual(values = c("External" = 0.6, "Internal" = 1.0)) +
    
    # Draw nodes
    ggraph::geom_node_point(
      aes(size = category, fill = category), 
      shape = 21, 
      color = "black" # border
    ) +
    # Define node styles
    scale_size_manual(values = c("Set Member" = 4, "Neighbor" = 0.1)) +
    scale_fill_manual(values = c("Set Member" = "#FFA500", "Neighbor" = "#DDDDDD")) +
    
    # Theme
    theme(legend.position="none") +
    coord_cartesian(clip = "off") + 
    theme(plot.margin = unit(c(1, 1, 1, 1), "cm"))
}

plot_urchin_graph <- function(ecp, core_set) {
  
  if (!requireNamespace("igraph", quietly = TRUE) || !requireNamespace("ggraph", quietly = TRUE)) {
    message("Packages 'igraph' and 'ggraph' are required to plot the graph.")
    return(invisible(NULL))
  }
  
  # --- 1. Identify Nodes and Edges ---
  core_nodes <- intersect(core_set, ecp@names)
  if (length(core_nodes) == 0) return(invisible(NULL))

  adj_subset <- ecp@adj[core_nodes]
  
  internal_edges_list <- list()
  external_stubs_list <- list()
  stub_nodes <- c()

  for (node in core_nodes) {
    all_neighbors <- adj_subset[[node]]
    internal_neighbors <- intersect(all_neighbors, core_nodes)
    external_neighbors <- setdiff(all_neighbors, core_nodes)
    
    n_external <- length(external_neighbors)
    
    # Create internal edges (only one way to avoid duplication)
    if (length(internal_neighbors) > 0) {
      # Only add edges to nodes that are "after" the current one to avoid A-B and B-A
      internal_neighbors_filtered <- internal_neighbors[internal_neighbors > node]
      if(length(internal_neighbors_filtered) > 0) {
        internal_edges_list <- c(internal_edges_list, list(data.table(
          from = node,
          to = internal_neighbors_filtered
        )))
      }
    }
    
    # Create "stub" nodes and edges for external neighbors
    if (n_external > 0) {
      new_stub_nodes <- paste0(node, "_stub_", 1:n_external)
      stub_nodes <- c(stub_nodes, new_stub_nodes)
      
      external_stubs_list <- c(external_stubs_list, list(data.table(
        from = node,
        to = new_stub_nodes
      )))
    }
  }

  # --- 2. Create Node Metadata ---
  nodes_df <- data.frame(
    name = c(core_nodes, stub_nodes),
    category = c(rep("Core Set", length(core_nodes)), 
                 rep("Spike", length(stub_nodes))),
    stringsAsFactors = FALSE
  )

  # --- 3. Create Edge Metadata ---
  internal_edges <- rbindlist(internal_edges_list)
  external_edges <- rbindlist(external_stubs_list)
  
  if (nrow(internal_edges) > 0) internal_edges[, type := "Internal"]
  if (nrow(external_edges) > 0) external_edges[, type := "External"]
  
  all_edges_df <- rbind(internal_edges, external_edges)
  
  # --- 4. Create igraph Object ---
  g <- igraph::graph_from_data_frame(all_edges_df, directed = FALSE, vertices = nodes_df)
  
  ggraph(g) +
    # Draw edges
    geom_edge_link(
      aes(width = type, alpha = type, color = type)
    ) +
    scale_edge_width_manual(values = c("External" = 4, "Internal" = 6)) +
    scale_edge_alpha_manual(values = c("External" = 0.7, "Internal" = 1.0)) +
    scale_edge_color_manual(values = c("External" = "darkgrey", "Internal" = "darkgreen")) +
    
    
    # Draw nodes
    geom_node_point(
      aes(size = category, fill = category, color = category), 
      shape = 21,
      stroke = 4
    ) +
    scale_size_manual(values = c("Core Set" = 26, "Spike" = 0.01)) +
    scale_fill_manual(values = c("Core Set" = "#FFC000", "Spike" = "white")) +
    scale_color_manual(values = c("Core Set" = "black", "Spike" = "white")) +
    coord_cartesian(clip = "off") + 
    theme_void() +
    theme(
      legend.position = "none",
      plot.margin = unit(c(1, 1, 1, 1), "cm")
    )
}
```

```{css, echo=FALSE}
body {
  font-size: 100%;
  line-height: 1.6;
  max-width: 940px;
  margin: 0 auto;
  padding: 20px;
}
img {
  border: none;
}
```
# Introduction
The `EdgeCount` package provides a statistical framework for enrichment analysis based on graph connectedness. It is designed for large, sparse graphs, such as protein interaction networks, social networks, or term-element membership graphs (bipartite graphs).

The package's methodology combines the Random Graph with Given Expected Degrees (RGGED) model [[1]](#ref1) with analytical expressions for connectedness effect size and p-value [[2]](#ref2). 

It also introduces Vertex Set Enrichment Analysis (VSEA), a permutation-based method similar to GSEA [[3]](#ref3). VSEA uniquely accounts for the degree of each element (i.e., the number of terms an element belongs to).

The primary functions allow users to:

- Evaluate a graph's suitability for some of the EdgeCount statistical models.
- Test for internal connectedness within one or more sets.
- Test for connectedness between pairs of sets.
- Score a collection of terms against a single set or a data table of element sets.
- Perform Vertex Set Enrichment Analysis (VSEA) on a ranked list of elements.
- Summarize VSEA results with Normalized Enrichment Scores (NES) and FDR q-values.

<br>

# Installation

The `EdgeCount` package is currently available on GitHub. You can install it using the `devtools` package. If you do not have `devtools` installed, you can install it from CRAN.

```{r installation, eval=FALSE}
# install.packages("devtools")
devtools::install_github("JoelRMath/EdgeCount")
```

<br>

# Basic Usage

This section provides an overview of the main use cases for the package.

## Creating a Graph

The core objects for any analysis are `ECGraph`, which holds the graph structure, and `ECProb`, which adds statistical model parameters and methods.

The first step is to create these objects from an edge list. The `ECGraph()` constructor is flexible and can accept this list as a `data.frame`, a file path, or an `igraph` object.

Here is an example for a graph of order 5 (number of vertices) and size 5 (number of edges).


```{r create_ECGraph_ECProb, message=FALSE, warning=FALSE}
library(EdgeCount)
library(data.table)
library(igraph)

# Data frame of edges 
edge_df <- data.frame(
  p1 = c("A", "B", "C", "D", "E"),
  p2 = c("B", "C", "A", "E", "A"),
  stringsAsFactors = FALSE
)

# ECGraph and ECProb objects from the edge list
ecg <- ECGraph(edge_df)
ecp <- ECProb(ecg)

# Same from a file
temp_file <- tempfile(fileext = ".tsv")
write.table(edge_df, file = temp_file, sep = "\t", row.names = FALSE, quote = FALSE)
ecg_from_file <- ECGraph(temp_file, col1 = "p1", col2 = "p2", header = TRUE, sep = "\t")
ecp_from_file <- ECProb(ecg_from_file)
unlink(temp_file)

# Same from an igraph object
igraph_obj <- igraph::graph_from_data_frame(edge_df, directed = FALSE)
ecg_from_igraph <- ECGraph(igraph_obj)
ecp_from_igraph <- ECProb(ecg_from_igraph)

# ECProb structure
str(ecp)
```

The S4 class `ECProb` inherits the slots of `ECGraph`:

 - `@names`: vertices names, as a chr vector
 - `@adj`: adjacency list, as a named list linking each vertex to its neighbors
 - `@degrees`: vertex degrees, as a named list
 
`ECProb` has the additional slots `@graph_size`, `@graph_order` and `@average_degree`, as well as statistical methods and functions that will be presented later. In order to properly use these functions, it is important to first check the suitability of the underlying `ECGraph` for some of them.

## Suitability for fast approximations

Even though it is not based on a large `ECGraph`, the previous example of `ECProb` object meets the criteria for fast approximations.

```{r plot_good_graph, message=FALSE, warning=FALSE, echo=FALSE, out.width="20%", fig.width=8, fig.height = 8, fig.align = "center"}

library(igraph)
library(ggraph)

# 1. Create the igraph object from the edge list
edge_df <- data.frame(
  p1 = c("A", "B", "C", "D", "E"),
  p2 = c("B", "C", "A", "E", "A"),
  stringsAsFactors = FALSE
)
g <- graph_from_data_frame(edge_df, directed = FALSE)

# 2. Create the plot
set.seed(1)
ggraph(g) +
    geom_edge_link(color = "darkgreen", width = 2) +
    geom_node_point(shape = 21, color = "black", size = 40, stroke = 1, fill="orange") +
    geom_node_text(aes(label = name), color = "black", size =20) +
    theme_void() +
    coord_cartesian(clip = "off") + 
    theme(plot.margin = unit(c(1, 1, 1, 1), "cm"))
```

This is verified by using method `summarize_suitability_fast` of `ECProb`.

```{r suitability_sparse}
suitability_stats <- summarize_suitability_fast(ecp)
print(suitability_stats)
```

- `$pij_over_1` is the important criterion: the proportion of edges among all possible pairs $(i,j)$ of vertices that yield a connection probability $p_{ij}$ greater than 1 (hence unsuitable) under an approximate analytical framework of the RGGED. A value close to 0 means that fast approximation methods are appropriate.
- `$prop_problematic_vertices` is the proportion of vertices with degree greater than $\sqrt{2M}$, where $M$ is the graph size `ecp@graph_size`. Indeed, under the model for fast approximations, $p_{ij} = k_ik_j/(2M)$, where $k$ is a vertex degree; so any vertex with $k_i > \sqrt{2M}$ might yield  $p_{ij} > 1$ for connection to another problematic vertex $j$.
- `$summary_pij` is a quick summary of the distribution of connection probabilities over all pairs of vertices
- `$pij_distribution` is the full distribution of $p_{ij}$ values in the RGGED model with fast approximations.

Here is now an example of small graph that, even though not a complete graph, does not pass the suitability test.

```{r plot_bad_graph, message=FALSE, warning=FALSE, echo=FALSE, out.width="20%", fig.width=8, fig.height = 8, fig.align = "center"}

edge_df <- data.frame(
  p1 = c("A", "A", "A", "A", "A", "B", "B", "B", "B", "C", "C", "C"),
  p2 = c("B", "C", "D", "E", "F", "C", "G", "H", "I", "J", "K", "L"),
  stringsAsFactors = FALSE
  )
g <- graph_from_data_frame(edge_df, directed = FALSE)

set.seed(2)
ggraph(g) +
    geom_edge_link(color = "darkgreen", width = 2) +
    geom_node_point(shape = 21, color = "black", size = 40, stroke = 1, fill="orange") +
    geom_node_text(aes(label = name), color = "black", size =20) +
    theme_void() +
    coord_cartesian(clip = "off") + 
    theme(plot.margin = unit(c(1, 1, 1, 1), "cm"))
```

```{r suitability_non_sparse}
edge_df <- data.frame(
  p1 = c("A", "A", "A", "A", "A", "B", "B", "B", "B", "C", "C", "C"),
  p2 = c("B", "C", "D", "E", "F", "C", "G", "H", "I", "J", "K", "L"),
  stringsAsFactors = FALSE
)
ecp <- ECProb(ECGraph(edge_df))
suitability_stats <- summarize_suitability_fast(ecp)
print(suitability_stats)
```


The `pij_over_1` value is 4.5%. This indicates that some vertex pairs have an approximate edge probability greater than or equal to 1, making the fast approximation methods unsuitable. For such a graph, other methods are available in the package.

In all rigor the fast approximation methods based on $p_{ij} = k_ik_j/(2M)$ are valid when $p_{ij} \ll 1$, otherwise $p_{ij} = 1-\exp(-k_ik_j/(2M))$ or $p_{ij} = \min(1, k_ik_j/(2M))$ need to be utilized, and the resulting methods are slower. 

**However, the fast approximation methods are conservative**:

In all rigor the fast approximation methods based on $p_{ij} = k_ik_j/(2M)$ are valid when $p_{ij} \ll 1$, otherwise $p_{ij} = 1-\exp(-k_ik_j/(2M))$ or $p_{ij} = \min(1, k_ik_j/(2M))$ need to be utilized, and the resulting methods are slower. **However**, by overestimating $p_{ij}$ values, the fast approximation methods are **conservative** for analysis:

- They overestimate $p_{ij}$ values.
- They underestimate effect sizes.
- They overestimate p-values.

**That is, fast approximation methods cannnot yield false-positive results**.

The package comes with object `sample_ecg`, an example of real large sparse graph based on the human network of the STRING database (version 12.0)[[4]](#ref4),[[5]](#ref5). The network is limited to protein interactions having a confidence score greater than or equal to 900. Proteins IDs have been replaced with their mapped Gene Entrez IDs. 

The package comes with object `sample_ecg`, an example of real large sparse graph based on the human network of the STRING database (version 12.0)[[4]](#ref4),[[5]](#ref5).

```{r sample_ecg}
data("sample_ecg")
data("sample_ects")
sample_ecp <- ECProb(sample_ecg)
str(sample_ecp, list.len = 6)
```
This network is suitable for fast approximations not only because `pij_over_1` = 0, but also because the majority of $p_{ij}$ values are close to $4.10^{-4}$.

This network is quite large and sparse.

```{r sample_ecg_suitability}
summarize_suitability_fast(sample_ecp)
```

It is suitable for fast approximations not only because `pij_over_1` = 0, but also because the majority of $p_{ij}$ values are close to $4.10^{-4}$. Object `sample_ecp` will be utilized next to provide examples of `ECProb` uses.

Object `sample_ecp` will be utilized next to provide examples of `ECProb` uses.

## Analysis of Interaction Networks (ECProb)

This workflow is for analyzing connectedness within an undirected, unipartite graph.

### Test for *within-set* connectedness

This test answers the question: "Are the vertices of a given set $S$ more connected to each other than expected by *chance*?" The concept of *chance* is here based on the RGGED: the distribution of the number of edges within set $S$ over all possible realizations of the RGGED, which in average preserve degrees of all vertices. Using such a null model that accounts for individual vertex degrees is statistically important.

```{r try_within_examples, echo=FALSE}
# line <- 48
line <- 48
df <- read.table("../data-raw/res/within_examples_selection.tsv", 
                 header = TRUE,
                 sep = "\t",
                 quote = "",
                 stringsAsFactors = FALSE)
set1_id <- df$set_id_at_min[line]
set2_id <- df$set_id_at_max[line]
```

```{r ecprob_within_example_1_urchin, message=FALSE, warning=FALSE, echo=FALSE, out.width="30%", fig.width=12, fig.height = 8, fig.align = "center"}
set1_id <- "GO:0032395"
set1_name <- "MHC class II receptor activity"
set1 <- get_neighbors(sample_ects@ecprob, set1_id)
plot_urchin_graph(sample_ecp, set1)
```


```{r ecprob_within_example_2_urchin, message=FALSE, warning=FALSE, echo=FALSE, out.width="30%", fig.width=12, fig.height = 8, fig.align = "center"}
set2_id <- "GO:0006729"
set2_name <- "tetrahydrobiopterin biosynthetic process"
set2 <- get_neighbors(sample_ects@ecprob, set2_id)
plot_urchin_graph(sample_ecp, set2)
```

### Test for *between-set* connectedness

This analysis answers the question: "Are my two sets more connected to each other than expected by chance?"

```{r ecprob_between_example}
# (Continuing from the example above)

# 1. Define two sets of interest
set_1 <- c("A", "B")
set_2 <- c("D", "E")

# 2. Calculate statistics
#    - First, get the observed edge count
obs_ec <- get_edge_count_between(ecp, set_1, set_2)
print(paste("Observed edges:", obs_ec))

#    - Then, get the full statistics
stats_between <- edge_count_statistics(
  ecp,
  set1 = set_1,
  set2 = set_2,
  observed_edge_count = obs_ec,
  lambda_method = "fast"
)
print(stats_between)
```

## Analysis of Bipartite Graphs (ECTermScoring)

This workflow is for analyzing term-element membership graphs (e.g., GO terms to genes). The main object for this is `ECTermScoring`. This is the foundation for Vertex Set Enrichment Analysis (VSEA).

```{r ects_example, eval=FALSE}
# This example is commented out to speed up the vignette build.
# You can run it manually in your console.

# 1. Load the sample ECTermScoring object
data("sample_ects")

# 2. Create a ranked list of elements
#    (Here we use a random ranking for demonstration)
set.seed(123)
ranked_elements <- sample(sample_ects@elements)
element_ranks <- setNames(seq_along(ranked_elements), ranked_elements)

# 3. Run the full VSEA analysis
#    (We use a small number of permutations for a quick example)
vsea_results <- run_vsea_analysis(
  sample_ects,
  element_ranks,
  n_permutations = 100,
  seed = 123
)

# 4. View the top results for positive enrichment
print(head(vsea_results$max_score_summary))
```

<br>

# References

<a id="ref1">(1.)</a> Chung & Lu (2002). The average distances in random graphs with given expected degrees. *PNAS*, 99 (25) 15879-1582. 
<a href="https://www.pnas.org/doi/10.1073/pnas.252631999" target="_blank">[link]</a>

<a id="ref2">(2.)</a> Pradines, J.R., et al. (2005). Analyzing Protein Lists with Large Networks: Edge-Count Probabilities in Random Graphs with Given Expected Degrees. *Journal of Computational Biology*, 12(2), 113-28.<a href="https://www.liebertpub.com/doi/10.1089/cmb.2005.12.113" target="_blank">[link]</a>


<a id="ref3">(3.)</a> Subramanian, A., et al. (2005). Gene set enrichment analysis: a knowledge-based approach for interpreting genome-wide expression profiles. *PNAS*, 102(43), 15545-50. <a href="https://www.pnas.org/doi/10.1073/pnas.0506580102" target="_blank">[link]</a>

<a id="ref4">(4.)</a> Szklarczyk D, Kirsch R, Koutrouli M, Nastou K, Mehryary F, Hachilif R, Annika GL, Fang T, Doncheva NT, Pyysalo S, Bork P, Jensen LJ, von Mering C. The STRING database in 2023: protein–protein association networks and functional enrichment analyses for any sequenced genome of interest. Nucleic Acids Res. 2023 Jan 6;51(D1):D638-646 <a href="https://pubmed.ncbi.nlm.nih.gov/36370105/" target= "_blank">[link]</a>

<a id="ref5">(5.)</a> Object `sample_ecg` is a subset of the human STRING network (version 12.0) with interaction scores of 900 or higher. Vertex IDs are Entrez Gene IDs [[8]](#ref8). The network has been trimmed to create a smaller, more manageable example for the package's vignettes and tests. Source: <a href="https://string-db.org/" target= "_blank">[string-db.org]</a>. Data from STRING is licensed under <a href="https://creativecommons.org/licenses/by/4.0/" target= "_blank">[CC BY 4.0]</a>.

<a id="ref6">(6.)</a> Ashburner et al. (2000). Gene ontology: tool for the unification of biology. *Nat Genet*, 25(1), 25-9. <a href="https://www.nature.com/articles/ng0500_25" target="_blank">[link]</a>

<a id="ref7">(7.)</a> The Gene Ontology Consortium. (2023). The Gene Ontology knowledgebase in 2023. *Genetics*, 224(1). <a href="https://doi.org/10.1093/genetics/iyad031" target="_blank">[link]</a>

<a id="ref8">(8.)</a> Carlson, M. (2019). org.Hs.eg.db: Genome wide annotation for Human. *R package*.

<a id="ref9">(9.)</a> Carlson, M. (2019). GO.db: A set of annotation maps describing the entire Gene Ontology. *R package*.

<a id="ref10">(10.)</a> `sample_ects` is derived from Gene Ontology (GO) term-to-gene mappings [[6, 7, 9]](#ref6). Element IDs are Entrez Gene IDs [[8]](#ref8). Source: <a href="http://geneontology.org/" target= "_blank">[geneontology.org]</a>. GO data is licensed under <a href="https://creativecommons.org/licenses/by/4.0/" target= "_blank">[CC BY 4.0]</a>.
