---
title: "EdgeCount Package Tutorial"  
author: "Joël R. Pradines"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
    number_sections: true
bibliography: references.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{EdgeCount Package Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{css, echo=FALSE}
/* 1. Prepare the container to handle absolute positioning */
div.csl-entry {
  position: relative;
  z-index: 1; /* Ensures the text stays on top of the highlight */
}

/* 2. Create the Highlight using a Pseudo-Element */
/* instead of modifying the element itself */
div.csl-entry:target::before {
  content: "";
  position: absolute;
  
  /* Sizing the box */
  top: -3px;
  bottom: -3px;
  right: -5px;
  /* This extends the yellow box to the left to cover the hanging indent */
  /* -2em is usually safe for standard bibliographies (which use 1.5em indent) */
  left: -2em; 
  
  /* Styling */
  background-color: #ffeb3b;
  border-radius: 4px;
  outline: 1px solid #fbc02d;
  
  /* Layering: Put it BEHIND the text */
  z-index: -1; 
}
div:target > h1,
div:target > h2,
div:target > h3,
div:target > h4,
div:target > h5,
div:target > h6,
section:target > h1,
section:target > h2,
section:target > h3 {
  background-color: #ffeb3b;
  transition: background-color 0.8s ease;
  display: inline-block;
  padding: 2px 8px;
  border-radius: 4px;
}
```

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
library(EdgeCount)
library(data.table)
library(patchwork)

plot_graph <- function(g){
  ggraph(g) +
    geom_edge_link(color = "blue", width = 3) +
    geom_node_point(shape = 21, color = "black", size = 40, stroke = 1, fill="orange") +
    geom_node_text(aes(label = name), color = "black", size =20) +
    theme_void() +
    coord_cartesian(clip = "off") + 
    theme(plot.margin = unit(c(1, 1, 1, 1), "cm"))
}

plot_urchin_graph <- function(ecp, core_set, symbol_lookup = NULL) {
  
  if (!requireNamespace("igraph", quietly = TRUE) || !requireNamespace("ggraph", quietly = TRUE)) {
    message("Packages 'igraph' and 'ggraph' are required to plot the graph.")
    return(invisible(NULL))
  }
  
  # --- 1. Identify Nodes and Edges ---
  core_nodes <- intersect(core_set, ecp@names)
  if (length(core_nodes) == 0) return(invisible(NULL))

  adj_subset <- ecp@adj[core_nodes]
  
  internal_edges_list <- list()
  external_stubs_list <- list()
  stub_nodes <- c()

  for (node in core_nodes) {
    all_neighbors <- adj_subset[[node]]
    internal_neighbors <- intersect(all_neighbors, core_nodes)
    external_neighbors <- setdiff(all_neighbors, core_nodes)
    
    n_external <- length(external_neighbors)
    
    # Create internal edges (only one way to avoid duplication)
    if (length(internal_neighbors) > 0) {
      # Only add edges to nodes that are "after" the current one to avoid A-B and B-A
      internal_neighbors_filtered <- internal_neighbors[internal_neighbors > node]
      if(length(internal_neighbors_filtered) > 0) {
        internal_edges_list <- c(internal_edges_list, list(data.table(
          from = node,
          to = internal_neighbors_filtered
        )))
      }
    }
    
    # Create "stub" nodes and edges for external neighbors
    if (n_external > 0) {
      new_stub_nodes <- paste0(node, "_stub_", 1:n_external)
      stub_nodes <- c(stub_nodes, new_stub_nodes)
      
      external_stubs_list <- c(external_stubs_list, list(data.table(
        from = node,
        to = new_stub_nodes
      )))
    }
  }

  # --- 2. Create Node Metadata ---
  nodes_df <- data.frame(
    name = c(core_nodes, stub_nodes),
    category = c(rep("Core Set", length(core_nodes)), 
                 rep("Spike", length(stub_nodes))),
    stringsAsFactors = FALSE
  )
  
  nodes_df$label <- NA_character_ 
  
  if (!is.null(symbol_lookup)) {
    # Get the indices for the core nodes
    core_indices <- which(nodes_df$category == "Core Set")
    core_node_ids <- nodes_df$name[core_indices]
    
    symbols <- symbol_lookup[core_node_ids]
    # Fallback to the original ID if a symbol is not found (NA)
    nodes_df$label[core_indices] <- ifelse(is.na(symbols), core_node_ids, symbols)
  }

  # --- 3. Create Edge Metadata ---
  internal_edges <- rbindlist(internal_edges_list)
  external_edges <- rbindlist(external_stubs_list)
  
  if (nrow(internal_edges) > 0) internal_edges[, type := "Internal"]
  if (nrow(external_edges) > 0) external_edges[, type := "External"]
  
  all_edges_df <- rbind(internal_edges, external_edges)
  
  # --- 4. Create igraph Object ---
  g <- igraph::graph_from_data_frame(all_edges_df, directed = FALSE, vertices = nodes_df)
  
  ggraph(g) +
    # Draw edges
    geom_edge_link(
      aes(width = type, alpha = type, color = type)
    ) +
    scale_edge_width_manual(values = c("External" = 4, "Internal" = 6)) +
    scale_edge_alpha_manual(values = c("External" = 0.7, "Internal" = 1.0)) +
    scale_edge_color_manual(values = c("External" = "lightblue", "Internal" = "blue")) +
    
    
    # Draw nodes
    geom_node_point(
      aes(size = category, fill = category, color = category), 
      shape = 21,
      stroke = 4
    ) +
    scale_size_manual(values = c("Core Set" = 26, "Spike" = 0.01)) +
    scale_fill_manual(values = c("Core Set" = "orange", "Spike" = "white")) +
    scale_color_manual(values = c("Core Set" = "black", "Spike" = "white")) +
    
    geom_node_text(aes(label = label), # 'label' column now has NAs for spikes
                   repel = TRUE, 
                   color = "black",
                   bg.color = NA,
                   size = 15) +
    
    coord_cartesian(clip = "off") + 
    theme_void() +
    theme(
      legend.position = "none",
      plot.margin = unit(c(1, 1, 1, 1), "cm")
    )
}

plot_between_comparison <- function(ecp, core_set_1, core_set_2, symbol_lookup = NULL) {
  
  if (!requireNamespace("igraph", quietly = TRUE) || !requireNamespace("ggraph", quietly = TRUE)) {
    message("Packages 'igraph' and 'ggraph' are required to plot the graph.")
    return(invisible(NULL))
  }
  
  # --- 1. Identify Node Sets (Disjoint Only) ---
  valid_set_1 <- intersect(core_set_1, ecp@names)
  valid_set_2 <- intersect(core_set_2, ecp@names)
  
  nodes_set_1_disjoint <- setdiff(valid_set_1, valid_set_2)
  nodes_set_2_disjoint <- setdiff(valid_set_2, valid_set_1)
  
  all_core_nodes <- c(nodes_set_1_disjoint, nodes_set_2_disjoint)
  
  if (length(all_core_nodes) == 0) return(invisible(NULL))

  # --- 2. Initialize Edge Lists ---
  intra_edges_list <- list()
  inter_edges_list <- list()

  # --- 3. Loop over all core nodes to categorize edges ---
  for (node in all_core_nodes) {
    all_neighbors <- ecp@adj[[node]]
    
    neighbors_in_set_1 <- intersect(all_neighbors, nodes_set_1_disjoint)
    neighbors_in_set_2 <- intersect(all_neighbors, nodes_set_2_disjoint)
    
    # --- Edge Logic ---
    if (node %in% nodes_set_1_disjoint) {
      # Add Intra-Set 1 edges
      if (length(neighbors_in_set_1) > 0) {
        internal_neighbors_filtered <- neighbors_in_set_1[neighbors_in_set_1 > node]
        if (length(internal_neighbors_filtered) > 0) {
          intra_edges_list <- c(intra_edges_list, list(data.table(
            from = node, to = internal_neighbors_filtered, type = "Intra-set", 
            layout_weight = 10  # <-- HIGH weight to keep cluster tight
          )))
        }
      }
      # Add Inter-Set edges (from Set 1 to Set 2)
      if (length(neighbors_in_set_2) > 0) {
        inter_edges_list <- c(inter_edges_list, list(data.table(
          from = node, to = neighbors_in_set_2, type = "Inter-set", 
          layout_weight = 1   # <-- LOW weight to allow separation
        )))
      }
    }
    
    # Add Intra-Set 2 edges (looping only from Set 2 nodes)
    if (node %in% nodes_set_2_disjoint) {
      if (length(neighbors_in_set_2) > 0) {
        internal_neighbors_filtered <- neighbors_in_set_2[neighbors_in_set_2 > node]
        if (length(internal_neighbors_filtered) > 0) {
          intra_edges_list <- c(intra_edges_list, list(data.table(
            from = node, to = internal_neighbors_filtered, type = "Intra-set", 
            layout_weight = 10  # <-- HIGH weight to keep cluster tight
          )))
        }
      }
    }
  }

  # --- 4. Create Node Metadata ---
  nodes_df <- data.frame(
    name = c(nodes_set_1_disjoint, nodes_set_2_disjoint),
    category = c(rep("Set 1", length(nodes_set_1_disjoint)), 
                 rep("Set 2", length(nodes_set_2_disjoint))),
    stringsAsFactors = FALSE
  )
  
  nodes_df$label <- NA_character_ 
  if (!is.null(symbol_lookup)) {
    core_indices <- which(nodes_df$category %in% c("Set 1", "Set 2"))
    core_node_ids <- nodes_df$name[core_indices]
    
    symbols <- symbol_lookup[core_node_ids]
    nodes_df$label[core_indices] <- ifelse(is.na(symbols), core_node_ids, symbols)
  }

  # --- 5. Create Final Edge Table ---
  all_edges_df <- rbindlist(c(intra_edges_list, inter_edges_list))
  
  # --- 6. Create igraph Object ---
  g <- igraph::graph_from_data_frame(all_edges_df, directed = FALSE, vertices = nodes_df)
  
  # --- 7. Plot with ggraph ---
  # Use the 'layout_weight' attribute to drive the 'fr' layout
  layout <- create_layout(g, layout = 'fr', weights = E(g)$layout_weight)
  
  ggraph(layout) +
    # Draw edges
    geom_edge_link(
      aes(width = type, alpha = type, color = type)
    ) +
    # Define *visual* edge styles (Inter-set is thick, Intra-set is thin)
    scale_edge_width_manual(values = c("Inter-set" = 4.0, "Intra-set" = 1)) +
    scale_edge_alpha_manual(values = c("Inter-set" = 1.0, "Intra-set" = 1)) +
    scale_edge_color_manual(values = c("Inter-set" = "blue", "Intra-set" = "black")) +
    
    # Draw nodes
    geom_node_point(
      aes(fill = category, color = category), 
      shape = 21,
      size = 34,
      stroke = 2
    ) +
    scale_fill_manual(values = c("Set 1" = "orange", "Set 2" = "green")) +
    scale_color_manual(values = c("Set 1" = "black", "Set 2" = "black")) +
    
    # Add labels
    geom_node_text(aes(label = label),
                   repel = TRUE, 
                   color = "black",
                   bg.color = NA,
                   size = 12) +
    
    coord_cartesian(clip = "off") + 
    theme_void() +
    theme(
      legend.position = "none",
      plot.margin = unit(c(1, 1, 1, 1), "cm")
    )
}

plot_bipartite <- function(term_element_df, node_sz){
  
nodes_df <- data.frame(
  name = unique(c(term_element_df$Term, term_element_df$Element)),
  stringsAsFactors = FALSE
)

nodes_df$is_element <- nodes_df$name %in% term_element_df$Element
nodes_df$category <- ifelse(nodes_df$is_element, "Element", "Term")

g_bipartite <- igraph::graph_from_data_frame(term_element_df, directed = FALSE, vertices = nodes_df)

V(g_bipartite)$type <- V(g_bipartite)$is_element

ggraph(g_bipartite, layout = 'bipartite') +
  
  geom_edge_link(color = "black", alpha = 1, width = 2) +
  
  geom_node_point(
    aes(shape = category, fill = category),
    color = "black",
    size = node_sz,
    stroke = 1
  ) +
  
  geom_node_text(aes(label = name), color = "black", size = 16) +
  
  scale_shape_manual(values = c("Term" = 22, "Element" = 21)) + 
  scale_fill_manual(values = c("Term" = "green", "Element" = "orange")) +
  
  # Flip coordinates to get Left-Right columns instead of Top-Bottom rows
  coord_cartesian(clip = "off") + 
  guides(
    color = guide_legend(override.aes = list(size = 1)), # For node color legend
  )+
  labs(fill = "Type", shape = "Type") +
  guides(
    # We must override FILL and SHAPE because those are the aesthetics in aes()
    # We set size = 8 so the legend icons are much smaller than the plot nodes (50)
    fill = guide_legend(override.aes = list(size = 12)),
    shape = guide_legend(override.aes = list(size = 12))
  ) +
    theme_void() +
    theme(
      legend.position = "bottom",
      plot.margin = unit(c(1.5, 1.5, 1.5, 1.5), "cm"),
      legend.box.spacing = unit(1.5, "cm"), 
      
      legend.text = element_text(size = 40), 
      legend.title = element_text(size = 40, face = "bold"),
      legend.key.size = unit(3, "cm")
    )
}
```

```{css, echo=FALSE}
body {
  font-size: 100%;
  line-height: 1.6;
  max-width: 940px;
  margin: 0 auto;
  padding: 20px;
}
img {
  border: none;
}
```
# Introduction
The `EdgeCount` package provides a statistical framework for enrichment analysis based on graph connectedness. It is designed for large, sparse graphs, such as protein interaction networks, social networks, or term-element membership graphs (bipartite graphs).

The package's methodology combines the Random Graph with Given Expected Degrees (RGGED) model [@chung2002] with analytical expressions for connectedness effect size and p-value [@pradines2005].

It also introduces Vertex Set Enrichment Analysis (VSEA), a permutation-based method similar to GSEA [@subramanian2005]. VSEA uniquely accounts for the degree of each element (i.e., the number of terms an element belongs to).

The primary functions allow users to:

- Evaluate a graph's suitability for some of the EdgeCount statistical models.
- Test for internal connectedness within one or more sets.
- Test for connectedness between pairs of sets.
- Score a collection of terms against a single set or a data table of element sets.
- Perform Vertex Set Enrichment Analysis (VSEA) on a ranked list of elements.
- Summarize VSEA results with Normalized Enrichment Scores (NES) and FDR q-values.

<br>

# Installation

The `EdgeCount` package is currently available on GitHub. It can be installed using the `devtools` package. 

```{r installation, eval=FALSE}
# install.packages("devtools")
devtools::install_github("JoelRMath/EdgeCount")
```

<br>

# Usage

This section provides an overview of the main use cases for the package.

## Creating a Graph

The core objects for any analysis are `ECGraph`, which holds the graph structure, and `ECProb`, which adds statistical model parameters and methods.

The first step is to create these objects from an edge list. The `ECGraph()` constructor is flexible and can accept this list as a `data.frame`, a file path, or an `igraph` object.

Here is an example for a graph of order 5 (number of vertices) and size 5 (number of edges).


```{r create_ECGraph_ECProb, message=FALSE, warning=FALSE}
library(EdgeCount)
library(data.table)
library(igraph)

# Data frame of edges 
edge_df <- data.frame(
  p1 = c("A", "B", "C", "D", "E"),
  p2 = c("B", "C", "A", "E", "A"),
  stringsAsFactors = FALSE
)

# ECGraph and ECProb objects from the edge list
ecg <- ECGraph(edge_df)
ecg
ecp <- ECProb(ecg)

# Same from a file
temp_file <- tempfile(fileext = ".tsv")
write.table(edge_df, file = temp_file, sep = "\t", row.names = FALSE, quote = FALSE)
ecg_from_file <- ECGraph(temp_file, col1 = "p1", col2 = "p2", header = TRUE, sep = "\t")
ecp_from_file <- ECProb(ecg_from_file)
unlink(temp_file)

# Same from an igraph object
igraph_obj <- igraph::graph_from_data_frame(edge_df, directed = FALSE)
ecg_from_igraph <- ECGraph(igraph_obj)
ecp_from_igraph <- ECProb(ecg_from_igraph)

# ECProb 
ecp
```

The S4 class `ECProb` inherits the slots of `ECGraph`:

 - `@names`: vertices names, as a chr vector
 - `@adj`: adjacency list, as a named list linking each vertex to its neighbors
 - `@degrees`: vertex degrees, as a named list
 
`ECProb` has the additional slots `@graph_size`, `@graph_order` and `@average_degree`, as well as statistical methods and functions that will be presented later. In order to properly use these functions, it is important to first check the suitability of the underlying `ECGraph` for some of them.

## Suitability for fast approximations

Even though it is not based on a large `ECGraph`, the previous example of `ECProb` object meets the criteria for fast approximations.

```{r plot_good_graph, message=FALSE, warning=FALSE, echo=FALSE, out.width="20%", fig.width=8, fig.height = 8, fig.align = "center"}

library(igraph)
library(ggraph)

# 1. Create the igraph object from the edge list
edge_df <- data.frame(
  p1 = c("A", "B", "C", "D", "E"),
  p2 = c("B", "C", "A", "E", "A"),
  stringsAsFactors = FALSE
)
g <- graph_from_data_frame(edge_df, directed = FALSE)

# 2. Create the plot
set.seed(1)
plot_graph(g)
```

This is verified by using method `summarize_suitability_fast` of `ECProb`.

```{r suitability_sparse}
suitability_stats <- summarize_suitability_fast(ecp)
suitability_stats
```

- `$pij_over_1` is the important criterion: the proportion of edges (among all possible pairs $(i,j)$ of vertices) that yield a connection probability $p_{ij}$ greater than 1 (hence unsuitable) under an approximate analytical framework of the RGGED. A value close to 0 means that fast approximation methods are appropriate.
- `$prop_problematic_vertices` is the proportion of vertices with degree greater than $\sqrt{2M}$, where $M$ is the graph size `ecp@graph_size`. Indeed, under the model for fast approximations, $p_{ij} = k_ik_j/(2M)$, where $k$ is a vertex degree; so any vertex with $k_i > \sqrt{2M}$ might yield  $p_{ij} > 1$ for connection to another vertex $j$.
- `$summary_pij` is a short summary of the distribution of connection probabilities over all pairs of vertices
- `$pij_distribution` is the full distribution of $p_{ij}$ values in the RGGED model with fast approximations.

Here is now a small graph that, even though not a complete graph, does not pass the suitability test.

```{r plot_bad_graph, message=FALSE, warning=FALSE, echo=FALSE, out.width="20%", fig.width=8, fig.height = 8, fig.align = "center"}

edge_df <- data.frame(
  p1 = c("A", "A", "A", "A", "A", "B", "B", "B", "B", "C", "C", "C"),
  p2 = c("B", "C", "D", "E", "F", "C", "G", "H", "I", "J", "K", "L"),
  stringsAsFactors = FALSE
  )
g <- graph_from_data_frame(edge_df, directed = FALSE)

set.seed(2)
plot_graph(g)
```

```{r suitability_non_sparse}
edge_df <- data.frame(
  p1 = c("A", "A", "A", "A", "A", "B", "B", "B", "B", "C", "C", "C"),
  p2 = c("B", "C", "D", "E", "F", "C", "G", "H", "I", "J", "K", "L"),
  stringsAsFactors = FALSE
)
ecp <- ECProb(ECGraph(edge_df))
suitability_stats <- summarize_suitability_fast(ecp)
suitability_stats
```


The `pij_over_1` value is 4.5%. This indicates that some vertex pairs have an approximate edge probability greater than or equal to 1, making the fast approximation methods unsuitable. For such a graph, other methods are available in the package.

The fast approximation methods are based on $p_{ij} = k_ik_j/(2M)$ and valid when $p_{ij} \ll 1$, otherwise $p_{ij} = 1-\exp(-k_ik_j/(2M))$ or $p_{ij} = \min(1, k_ik_j/(2M))$ need to be utilized, and the resulting methods are slower. 

This said, false-positive results cannot be obtained with fast approximation methods. Indeed,

- they overestimate $p_{ij}$ values,
- they underestimate effect sizes,
- they overestimate edge-count p-values.

In other words, using the fast approximation methods by default is a conservative approach to analysis.

The package comes with object `sample_ecg`, an example of real large sparse graph based on the human network of the STRING database (version 12.0)[@szklarczyk2023],[@sample_ecg_data]. The network is limited to protein interactions having a confidence score greater than or equal to 900. Proteins IDs have been replaced with their mapped Gene Entrez IDs. 


```{r sample_ecg}
data("sample_ecg")
sample_ecp <- ECProb(sample_ecg)
sample_ecp
```

This network is relatively large and it is sparse.

```{r sample_ecg_suitability}
system.time(suitability <- summarize_suitability_fast(sample_ecp))
suitability
```

Indeed, it is suitable for fast approximations not only because `pij_over_1` = 0, but also because the majority of $p_{ij}$ values are close to $4.10^{-4}$. 

Object `sample_ecp` is utilized next to provide examples of `ECProb` uses.

## Analysis with Interaction Networks (ECProb)

Example of connectedness analyses within an undirected, unipartite graph are now presented.

### Tests for *within-set* connectedness

This test answers the question: "Are the vertices of a given set $S$ more connected to each other than expected by *chance*?" The concept of *chance* is here based on the RGGED: the distribution of the number of edges within set $S$ over all possible realizations of the RGGED, which in average preserve degrees of all vertices. Using such a null model that accounts for individual vertex degrees is statistically important.

#### Basic method 

The package also comes with a sample bipartite network `sample_ects` that maps a trimmed version of GO categories to their human gene elements [@orgHsEg], [@goconsortium2023], [@sample_ects_data]. Object `sample_ects@ecprob` is utilized below to define with a term a set of elements/vertices which is then tested for its connectedness in interaction network `sample_ecp`.

```{r setA_in, message=FALSE, warning=FALSE}
# bipartite network (term-element relationships)
data("sample_ects")
# term descriptions
data("sample_term_lookup")
setA_id <- "GO:0032395"
setA_name <- sample_term_lookup[setA_id]
setA_name
# elements mapped to term (bipartite network sample_ects@ecprob)
setA <- get_neighbors(sample_ects@ecprob, setA_id) 
length(setA)
# observed edge count in the graph between elements (interaction network sample_ecg)
data("sample_ecg")
sample_ecp <- ECProb(sample_ecg)
observed_ec <- get_edge_count_in(sample_ecp, setA) 
# calculation of statistics
statsA <- edge_count_statistics(sample_ecp, 
                               set1 = setA, 
                               set2 = NULL, 
                               observed_edge_count = observed_ec,
                               lambda_method = "fast")
statsA
```

Method `edge_count_statistics` returns a list:

- `$observed_edge_count` $x$ is the observed number of edges, here the edge count within `setA`
- `max_possible_edges` $m$ is the maximum number of edges one could observe in the RGGED, here the number of pairs withing `setA`, i.e. $(7\times6)/2 = 21$
- `$lambda`, $\lambda$ is the expected number of observed edges within `setA` in the RGGED, that is, the average number of edges over all realizations of the RGGED
- `$p-value` $p_S$ is $\Pr\left( X \geq x\right)$, where $X$ is the random variable 'number of edges within `setA`' in the RGGED
- `$log2_Anscombe_ratio` $lAr$ is the observed *log effect size*
- `$size1` is the size of the input set reduced to its vertices represented in the `ECProb` object
- `$size2` is used only for between-sets connectedness. Here it is zero because argument `set2` is null

Calling $p_{ij}$ the probability of connecting two vertices $i$ and $j$ in the RGGED, $k$ the degree of a vertex, $M$ the graph size and $X$ the random variable 'number of edges within $S$ under the RGGED model', we have

$$p_{ij} = \min\left(1, \frac{k_ik_j}{2M}\right)$$
$$m = \frac{1}{2}\left| S \right|\left( \left| S\right|-1\right)$$
$$\lambda = \sum_{i < j, \in S} p_{ij}$$
$$p_S = \sum_{u=x}^m \Pr\left(X = u\right)$$
$$\Pr\left(X = u\right) = \frac{e^{-\lambda}}{\alpha}\frac{\lambda^u}{u!}$$
$$\alpha = e^{-\lambda}\sum_{v=0}^{m} \frac{\lambda^v}{v!}$$
$$lAr = \frac{1}{2} \log_2 \frac{x +3/8}{\lambda + 3/8}$$
Using $lAr$ as effect size is important because $X$ can have a truncated Poisson distribution of parameter $\lambda$ small enough to yield an asymmetric distribution.

With `lambda_method = "fast"` calculations are simplified

$$p_{ij} = \frac{k_ik_j}{2M} \quad \text{and} \quad \lambda = \frac{1}{4M} \left( \left(\sum_{i\in S} k_i\right)^2 - \sum_{i\in S} k_i^2\right)$$
Namely, time complexity is now $O\left( \left| S \right|\right)$ instead of $O\left( \left| S \right|^2\right)$.

To illustrate the importance of accounting for the degree of each vertex consider now the following `setB`.

```{r setB_in, message=FALSE, warning=FALSE}
# bipartite network (term-element relationships)
data("sample_ects")
setB_id <- "GO:0006729"
# term descriptions
data("sample_term_lookup")
setB_name <- sample_term_lookup[setB_id]
setB_name
# elements mapped to term (bipartite network sample_ects@ecprob)
setB <- get_neighbors(sample_ects@ecprob, setB_id) 
length(setB)
# observed edge count in the graph between elements (interaction network sample_ecg)
data("sample_ecg")
sample_ecp <- ECProb(sample_ecg)
observed_ec <- get_edge_count_in(sample_ecp, setB) 
# calculation of statistics
statsB <- edge_count_statistics(sample_ecp, 
                               set1 = setB, 
                               set2 = NULL, 
                               observed_edge_count = observed_ec,
                               lambda_method = "fast")
statsB
```

While `setA` and `setB` have the same size, and thus the same `max_possible_edges`, and the same `observed_edge-count`, `setB` has lesser connectedness than `setA`, as attested by a larger `p-value` and a smaller `log2_Anscombe_ratio`. This is due to `setB` having a larger `lambda` than `setA` because its vertices have larger degrees. This is illustrated below by plotting the subgraphs induced by the two sets while complementing visualization of their degrees in the entire network in the form of stubs.


```{r plot_urchin_comparison, message=FALSE, warning=FALSE, echo=FALSE, fig.width=25, fig.height=12, out.width="80%", fig.align = "center"}
library(patchwork)

tsz <- 35

# 1. Create Plot A and save it to a variable
p_A <- plot_urchin_graph(sample_ecp, setA, sample_gene_symbol_lookup) + 
  ggtitle(paste0("Set A: ", setA_id, "\n", setA_name, "\n",
                 "set_size = ", length(setA), 
                 ", observed_ec = ", statsA$observed_edge_count, "\n",
                 "p_value = ", signif(statsA$p_value, digits = 3), "\n",
                 "log2_Anscombe_ratio = ", signif(statsA$log2_Anscombe_ratio, digits = 3))) + 
  theme(plot.title = element_text(size = tsz))

# 2. Create Plot B and save it to a variable
p_B <- plot_urchin_graph(sample_ecp, setB, sample_gene_symbol_lookup) + 
  ggtitle(paste0("Set B: ", setB_id, "\n", setB_name, "\n",
                 "set_size = ", length(setB), 
                 ", observed_ec = ", statsB$observed_edge_count, "\n",
                 "p_value = ", signif(statsB$p_value, digits = 3), "\n",
                 "log2_Anscombe_ratio = ", signif(statsB$log2_Anscombe_ratio, digits = 3)))  + 
  theme(plot.title = element_text(size = tsz))

# 3
p_A + p_B
```

#### Vectorized method

The last examples illustrated how to test an individual set for its within connectedness. When dealing with multiple sets, it is better to utilize a vectorized method which has been optimized for speed, by using the `data.table` package [@datatable2025] and fast lambda approximation.

Method `calculate_in_stats_fast_vectorized(ecprob, sets_dt, set_membership_dt)` has the following arguments

- `ecprob`: an instance of `ECProb` (interaction network)
- `sets_dt`: a data table with a single column ("set_id") containing IDs of sets to test
- `set_membership_dt': a data table with "set_id" and "element" columns (bipartite network)

Below is the application of this method to the entire collection of sets  contained in `sample_ects` (GO categories [@orgHsEg],[@goconsortium2023], [@sample_ects_data]) and the interaction network `sample_ecg` (extracted from STRING [@szklarczyk2023],[@sample_ecg_data]).


```{r in_vectorized}
# interaction network and associated ECProb
data("sample_ecg")
ecprob <- ECProb(sample_ecg)
# bipartite graph: term-element relationships
data("sample_ects")
set_membership_dt <- as.data.table(to_dataframe(sample_ects))
setnames(set_membership_dt, c("term", "element"), c("set_id", "element"))
sets_to_test <- data.table(set_id = unique(set_membership_dt$set_id))
length(sets_to_test$set_id)
length(set_membership_dt$element)
# within conncectedness statistics for all sets
system.time(results <- calculate_in_stats_fast_vectorized(ecprob, sets_to_test, set_membership_dt))
# adding annotation (term names, not just ids)
data("sample_term_lookup")
lookup_dt <- data.table(
  set_id = names(sample_term_lookup),
  term_name = sample_term_lookup
)
setkey(results, set_id)
setkey(lookup_dt, set_id)
annotated_results <- lookup_dt[results, on = "set_id"]
setorder(annotated_results, -log2_Anscombe_ratio)
num_cols <- names(annotated_results)[sapply(annotated_results, is.numeric)]
annotated_results[, (num_cols) := lapply(.SD, signif, digits = 2), .SDcols = num_cols]

annotated_results
```

Note that `set_size` corresponds to that of the input set reduced to vertices represented in the `ECProb` object. Distributions of some of the statistics in the above returned data table are shown below.

```{r plot_distributions_in, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=8, fig.align="center", out.width="90%"}
library(ggplot2)
library(patchwork)

p1 <- ggplot(annotated_results, aes(x = log10(observed_edge_count + 1))) +
  geom_histogram(bins = 30, fill = "#0c4a6e", color = "white", alpha = 0.8) +
  ggtitle("Observed Edge Count") +
  xlab("log10(observed_edge_count + 1)") +
  theme_minimal(base_size = 14)

p2 <- ggplot(annotated_results[lambda >0 ], aes(x = log10(lambda))) +
  geom_histogram(bins = 30, fill = "#047857", color = "white", alpha = 0.8) +
  ggtitle("Expected Edge Count") +
  xlab("log10(lambda)") +
  theme_minimal(base_size = 14)

p3 <- ggplot(annotated_results, aes(x = p_value)) +
  geom_histogram(bins = 20, boundary = 0, fill = "#FF7518", color = "white", alpha = 0.8) +
  ggtitle("P-Value") +
  xlab("p_value") +
  theme_minimal(base_size = 14)

p4 <- ggplot(annotated_results, aes(x = log2_Anscombe_ratio)) +
  geom_histogram(bins = 40, fill = "#9a3412", color = "white", alpha = 0.8) +
  ggtitle("Effect Size") +
  xlab("log2_Anscombe_ratio") +
  theme_minimal(base_size = 14)

(p1 + p2) / (p3 + p4)
```

### Test for *between-sets* connectedness

This test answers the question: "Are two given vertex sets $S_1$ and $S_2$ more connected to each other than expected by *chance*?" The concept of *chance* is again based on the RGGED: the distribution of the number of edges between $S_1$ and $S_2$ over all possible realizations of the RGGED, a null model that accounts for individual vertex degrees. 

The 'between-sets' connectedness is a concept different from that of 'within-connectedness' for individual sets. Do two sets 'talk' to each other, independently of their own internal connectedness? That is, pairs of sets are first reduced to their specific parts and then these two disjoint parts are tested for connectedness with each other. 

#### Basic method

The basic method is again `edge_count_statistics`. Two sets $S_1$ and $S_2$ are passed as arguments, and argument `observed_edge_count` is now obtained with method `get_edge_count_between`.

Below is an example with two vertex sets represented in bipartite network `sample_ects` and tested for their 'between connectedness' with interaction network `sample_ecg` and its `ECProb` instance `sample_ecp`.

```{r between_basic, message=FALSE, warning=FALSE}
# bipartite network (term-element relationships)
data("sample_ects")
# selection of two terms and their descriptions
data("sample_term_lookup")
set1_id <- "GO:0000221"
set1_name <- sample_term_lookup[set1_id]
set1_name
set2_id <- "GO:0016471"
set2_name <- sample_term_lookup[set2_id]
set2_name
# elements mapped to terms (bipartite network sample_ects@ecprob)
set1 <- get_neighbors(sample_ects@ecprob, set1_id) 
set2 <- get_neighbors(sample_ects@ecprob, set2_id) 
# observed edge count in the graph between the two sets (interaction network sample_ecg)
data("sample_ecg")
sample_ecp <- ECProb(sample_ecg)
observed_ec <- get_edge_count_between(sample_ecp, set1, set2) 
# calculation of statistics
stats <- edge_count_statistics(sample_ecp, 
                               set1 = set1, 
                               set2 = set2, 
                               observed_edge_count = observed_ec,
                               lambda_method = "fast")
stats
```


Method `edge_count_statistics` returns a list:

- `$observed_edge_count` $x$ is the observed number of edges, here the edge count between the disjoint parts of `set1` and `set2`
- `max_possible_edges` $m$ is the maximum number of edges one could observe in the RGGED, here the number of possible edges between the disjoint parts of `set1` and `set2` reduced to vertices represented in the `ECProb` object, that is $5\times 6 = 30$.
- `$lambda`, $\lambda$ is the expected number of observed edges between disjoint sets `set1` and `set2` in the RGGED, that is, the average number of edges over all realizations of the RGGED
- `$p-value` $p_{S_1,S_2}$ is $\Pr\left( X \geq x\right)$, where $X$ is the random variable 'number of edges between `set1` and `set2`' in the RGGED
- `$log2_Anscombe_ratio` $lAr$ is the observed *log effect size*
- `$size1` and `size2`: sizes of the disjoint parts of the input sets after their reduction to vertices represented in the `ECProb` object

Calling $p_{ij}$ the probability of connecting two vertices $i$ and $j$ in the RGGED, $k$ the degree of a vertex, $M$ the graph size and $X$ the RGGED random variable 'observed number of edges between $S_1$ and $S_2$', we have

$$p_{ij} = \min\left(1, \frac{k_ik_j}{2M}\right)$$
$$m = \left| S_1 \right| \left| S_2\right|$$
$$\lambda = \sum_{i \in S_1, j \in S_2} p_{ij}$$
$$p_{S_1,S_2} = \sum_{u=x}^m \Pr\left(X = u\right)$$
$$\Pr\left(X = u\right) = \frac{e^{-\lambda}}{\alpha}\frac{\lambda^u}{u!}$$
$$\alpha = e^{-\lambda}\sum_{v=0}^{m} \frac{\lambda^v}{v!}$$
$$lAr = \frac{1}{2} \log_2 \frac{x +3/8}{\lambda + 3/8}$$
Using $lAr$ as effect size is important because $X$ can have a truncated Poisson distribution of parameter $\lambda$ small enough to yield an asymmetric distribution.

With `lambda_method = "fast"` calculations are simplified as follows

$$p_{ij} = \frac{k_ik_j}{2M} \quad \text{and} \quad \lambda = \frac{1}{2M} \left(\sum_{i\in S_1} k_i\right) \left( \sum_{j\in S_2} k_j \right)$$
That is, time complexity is now $O\left( \left| S_1 \right| + \left| S_2 \right| \right)$ instead of $O\left( \left| S_1 \right| \times \left| S_2 \right| \right)$.

As previously mentioned, using the RGGED model takes into account individual vertex degrees in the quantification of connectedness. To illustrate this, consider now the pair of `set3` (GO\:0071986) and `set4` (GO\:1990130), that yields the same doublet of set sizes ($\left\{ 5, 6\right\}$), and thus the same maximum number of 'between' edges (30), and the same observed number of edges (21). Because vertices of the two pairs have different degrees, the two pairs yield different estimations of connectedness:

```{r plot_between_comparison, echo=FALSE, fig.align="center", fig.height=12, fig.width=25, message=FALSE, warning=FALSE, out.width="90%"}
library(ggtext)
set1_id <- "GO:0000221"
set2_id <- "GO:0016471"
set3_id <- "GO:0071986"
set4_id <- "GO:1990130"
data("sample_ects")
data("sample_term_lookup")
set1_name <- sample_term_lookup[set1_id]
set2_name <- sample_term_lookup[set2_id]
set3_name <- sample_term_lookup[set3_id]
set4_name <- sample_term_lookup[set4_id]

set1 <- get_neighbors(sample_ects@ecprob, set1_id) 
set2 <- get_neighbors(sample_ects@ecprob, set2_id) 
observed_ec1 <- get_edge_count_between(sample_ecp, set1, set2) 
stats1 <- edge_count_statistics(sample_ecp, 
                               set1 = set1, 
                               set2 = set2, 
                               observed_edge_count = observed_ec1,
                               lambda_method = "fast")

set3 <- get_neighbors(sample_ects@ecprob, set3_id) 
set4 <- get_neighbors(sample_ects@ecprob, set4_id) 
observed_ec2 <- get_edge_count_between(sample_ecp, set3, set4) 
stats2 <- edge_count_statistics(sample_ecp, 
                               set1 = set3, 
                               set2 = set4, 
                               observed_edge_count = observed_ec2,
                               lambda_method = "fast")

data("sample_ecg")
sample_ecp <- ECProb(sample_ecg)

tsz <- 30

# 1. Create Plot A and save it to a variable
p_A <- plot_between_comparison(sample_ecp, set1, set2, sample_gene_symbol_lookup) + 
  ggtitle(paste0("<span style='color:#C76E00;'> Set 1: ", set1_id, ", (", stats1$size1, " elements) <br>",
                 set1_name, "</span><br>",
                 "<span style='color:#50C878;'> Set 2: ", set2_id, ", (", stats1$size2, " elements) <br>",
                 set2_name, "</span><br>",
                "<span style='color:blue;'> observed_edge_count = ", stats1$observed_edge_count, "</span><br> <br> ",
                "lambda = ", signif(stats1$lambda, digits=3), "<br>",
                "p_value = ", signif(stats1$p_value, digits=3), "<br>","log2_Anscombe_ratio = ", signif(stats1$log2_Anscombe_ratio, digits=3), "<br>")) + 
  theme(plot.title = element_markdown(size = tsz, hjust = 0), plot.margin = unit(c(1, 6, 1, 1), "cm") )

# 2. Create Plot B and save it to a variable
p_B <- plot_between_comparison(sample_ecp, set3, set4, sample_gene_symbol_lookup) + 
 ggtitle(paste0("<span style='color:#C76E00;'> Set 3: ", set3_id, ", (", stats2$size1," elements) <br>",
                set3_name, "</span><br>",
                 "<span style='color:#50C878;'> Set 4: ", set4_id,", (", stats2$size2, " elements) <br>",
                set4_name, "</span><br>",
                "<span style='color:blue;'> observed_edge_count = ", stats2$observed_edge_count, "</span><br> <br> ",
                "lambda = ", signif(stats2$lambda, digits=3), "<br>", 
                "p_value = ", signif(stats2$p_value, digits=3), "<br>",
                "log2_Anscombe_ratio = ", signif(stats2$log2_Anscombe_ratio, digits=3), "<br>"))  + 
  theme(plot.title = element_markdown(size = tsz, hjust = 0), plot.margin = unit(c(1, 1, 1, 6), "cm") )

# 3
p_A + p_B
```

#### Vectorized method

The last examples showed how to test a single pair of sets for their between connectedness in an interaction network. When dealing with multiple pairs, it is better to utilize a vectorized method which has been optimized for speed and takes a data table of pairs as argument. Note that this method relies on a fast lambda approximation. 

Method `calculate_between_stats_fast_vectorized` has the following arguments

- `object`: the `ECProb` object representing an interaction network
- `pairs_dt`: a data table of pairs of set identifiers ("set1_id", "set2_id")
- `set_membership_dt`: a data table representing set-element membership ("set_id", "element")

Below is the application of this method to all possible pairs of sets  contained in `sample_ects` (GO categories [@orgHsEg],[@goconsortium2023], [@sample_ects_data]) and the interaction network `sample_ecg` (extracted from STRING [@szklarczyk2023],[@sample_ecg_data]).

Numbers of terms and possible pairs are quite large.

```{r sample_ects_pairs}
data("sample_ects")
data("sample_ecg")
sample_ecp <- ECProb(sample_ecg)
number_of_terms <- length(sample_ects@terms)
number_of_terms
number_of_pairs <- number_of_terms*(number_of_terms-1)/2
number_of_pairs
```

Because `sample_ecp` is a sparse network, it is better to focus on pairs that are connected by at least one edge in the interaction network. Indeed, if a pair of sets has no interaction, its p-value is 1 and we can ignore it: 
$$p_{S_1,S_2} = 1 \quad \text{and} \quad \text{lAr } = \log_2 \frac{3/8}{\lambda+3/8} < 0.$$
The package contains `ECProb` method `get_disjoint_connected_pairs` that efficiently reduces a data table of set pairs to those pairs that are at least connected by one edge through their disjoint parts (the between connectedness is estimated on the disjoint parts). 

Below is the application to pairs of sets derived from `sample_ects`.

```{r pairs_sparse}
set_membership <- as.data.table(to_dataframe(sample_ects))
setnames(set_membership, c("term", "element"), c("set_id", "element"))
system.time(
  connected_pairs <- get_disjoint_connected_pairs(sample_ecp, 
                                                  set_membership,
                                                  sets = NULL)
)
setorder(connected_pairs, -observed_edges)
connected_pairs
nrow(connected_pairs)
signif(nrow(connected_pairs)/number_of_pairs, 2)
```
Note that method `get_disjoint_connected_pairs` has a third optional argument `sets` that can be utilized to focus on some of the sets represented in the second argument `set_membership`. The method rapidly returms a much smaller number of pairs (0.75%) than the full Cartesian product.

Next, `connected_pairs` are efficiently scored for their between connectedness with method `calculate_between_stats_fast_vectorized`. This method is fully vectorized by using the `data.table` package [@datatable2025] and fast lambda approximations.

```{r vectorized_between}
# fast calculation of between statistics for candidate pairs
set_membership_dt <- as.data.table(to_dataframe(sample_ects))
setnames(set_membership_dt, c("term", "element"), c("set_id", "element"))
system.time(
  results <- calculate_between_stats_fast_vectorized(sample_ecp, 
                                                   connected_pairs,
                                                   set_membership_dt)
)
setorder(results, -log2_Anscombe_ratio)
num_cols <- names(results)[sapply(results, is.numeric)]
results[, (num_cols) := lapply(.SD, signif, digits = 2), .SDcols = num_cols]
results
```

Distributions of some of the statistics obtained with the above example are shown below.

```{r plot_distributions_between, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=8, fig.align="center", out.width="90%"}
library(ggplot2)
library(patchwork)

p1 <- ggplot(results, aes(x = log10(observed_edges))) +
  geom_histogram(bins = 30, fill = "#0c4a6e", color = "white", alpha = 0.8) +
  ggtitle("Observed Edge Count (> 0)") +
  xlab("log10(observed_edge_count)") +
  theme_minimal(base_size = 14)

p2 <- ggplot(results, aes(x = log10(lambda))) +
  geom_histogram(bins = 30, fill = "#047857", color = "white", alpha = 0.8) +
  ggtitle("Expected Edge Count (> 0)") +
  xlab("log10(lambda)") +
  theme_minimal(base_size = 14)

p3 <- ggplot(results, aes(x = p_value)) +
  geom_histogram(bins = 30, boundary = 0, fill = "#FF7518", color = "white", alpha = 0.8) +
  ggtitle("P-Value (< 1)") +
  xlab("p_value") +
  theme_minimal(base_size = 14)

p4 <- ggplot(results, aes(x = log2_Anscombe_ratio)) +
  geom_histogram(bins = 30, fill = "#9a3412", color = "white", alpha = 0.8) +
  ggtitle("Effect Size") +
  xlab("log2_Anscombe_ratio") +
  theme_minimal(base_size = 14)

(p1 + p2) / (p3 + p4)
```

Remember that the above plots are restricted to pairs having at least one connection. Over all possible pairs in defined by the sets/terms of `sample_ects`, the proportion of p-values equal to 1 is actually 99.25%.

## Analysis with Bipartite Graphs

The Edge-Count statistical framework based on the RGGED and utilized for data analysis with an undirected graph represented by `ECGraph` and `ECProb` objects, can also be applied to the case of bipartite networks: a collection of sets (called `terms`) of vertices (called `elements`). In this network, edges exist only between vertices of different types. Namely, the network represents membership of elements in terms.


### The ECTermScoring class

S4 class `ECTermScoring` represents an undirected bipartite graph and some associated methods. It is built from class `ECProb` and has additional slots:

- `@ecprob`: represents an underlying `ECGraph` and its associated statistical methods
- `@elements`: a vector of vertex names in `ECProb`
- `@terms`: another vector of vertex names in `ECProb`, disjoint from `@elements`

Undirected edges run only bewteen a `term` and an `element`. That is, an `ECTermScoring` instance is a collection of terms, each having its own set of elements, though these sets can overlap between terms. The degree $k_t$ of a term is the number of elements it describes, and the degree $k_e$ of an element is the number of terms it is described by.

The argument of the `ECTermScoring` constructor can be either a data frame of bipartite edges or a file path.

```{r bipartite_plot, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=6, fig.align="center", out.width="30%"}
term_element_df <- data.frame(
  Term = c("T1", "T1", "T2", "T2", "T3"),
  Element = c("E1", "E2", "E2", "E3", "E4"),
  stringsAsFactors = FALSE
)
plot_bipartite(term_element_df, node_sz = 45)

```

```{r ects_constructor}
  term_element_df <- data.frame(
    Term = c("T1", "T1", "T2", "T2", "T3"),
    Element = c("E1", "E2", "E2", "E3", "E4"),
    stringsAsFactors = FALSE
  )
  ects <- ECTermScoring(term_element_df)
  
  temp_file <- tempfile(fileext = ".tsv")
  write.table(term_element_df, file = temp_file, sep = "\t", row.names = FALSE, quote = FALSE)
  ects <- ECTermScoring(temp_file)
  ects
  unlink(temp_file)
```

Note that the constructor will fail if a vertex appears both as term and element in the provided edge list.

### Suitability for fast approximations

The `@ecprob` slot of an `ECTermScoring` object allows for calculation of edge-count statistics on the underlying bipartite graph, including some based on fast lambda approximations. Because this graph is bipartite, any edge counting and associated statistics is done on edges between terms and elements. In particular, the method `summarize_suitability_fast`, used to check a graph for suitability to fast lambda approximations, is not optimal in the context of an `ECTermScoring` object. Instead, `ECTermScoring` contains method `summarize_suitability_bipartite` which is specifically designed for bipartite graphs.

Here is an example of small bipartite graph that is not optimal for fast lambda approximation.

```{r plot_inap_bip, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=6, fig.align="center", out.width="35%"}
term_element_df_p <- data.frame(
  Term = c("T1", "T1","T2", "T2", "T3", "T3", "T4","T4", "T4", "T4", "T4", "T4","T5", "T5", "T6", "T6"),
  Element = c("E1","E3","E2", "E3", "E2", "E3", "E1","E4", "E2", "E3", "E5", "E6", "E3", "E5", "E3", "E6"),
  stringsAsFactors = FALSE
)
plot_bipartite(term_element_df_p, node_sz = 40)
```

```{r suitability_bipartite}
term_element_df_p <- data.frame(
  Term = c("T1", "T1","T2", "T2", "T3", "T3", "T4","T4", "T4", "T4", "T4", "T4","T5", "T5", "T6", "T6"),
  Element = c("E1","E3","E2", "E3", "E2", "E3", "E1","E4", "E2", "E3", "E5", "E6", "E3", "E5", "E3", "E6"),
  stringsAsFactors = FALSE
)
ects <- ECTermScoring(term_element_df_p)
summarize_suitability_bipartite(ects)
```

The returned list has the following items:

- `$pij_over_1`: the proportion of edges (among all possible pairs (term $i$, element $j$)) that yield a connection probability $p_{ij}$ greater than 1 (hence unsuitable) under an approximate analytical framework of the RGGED. A value close to 0 means that fast approximation methods are appropriate.
- `$prop_problematic_terms` is the proportion of terms with degree greater than $\sqrt{2M}$, where $M$ is the graph size `ecp@graph_size`. Indeed, under the model for fast approximations, $p_{ij} = k_ik_j/(2M)$, where $k$ is a vertex degree; so any term with $k_i > \sqrt{2M}$ might yield  $p_{ij} > 1$ with an element.
- `$prop_problematic_elements` is the proportion of terms with degree greater than $\sqrt{2M}$. Note that `$prop_problematic_elements` > 0 does not imply `$prop_problematic_terms` > 0. Likewise, `$prop_problematic_terms` > 0 does not imply `$prop_problematic_elements` > 0.
- `$summary_pij` is a short summary of the distribution of connection probabilities over all possible term-element pairs 
- `$pij_distribution` is the full distribution of $p_{ij}$ values in the RGGED model with fast approximations.

This said, false-positive results cannot be obtained with fast approximation methods, because

- they overestimate $p_{ij}$ and $\lambda$ values,
- they underestimate effect sizes,
- they overestimate edge-count p-values.

In other words, using the fast approximation methods by default is a conservative approach to analysis.


The EdgeCount package comes with an example of relatively large bipartite graph `sample_ects`, a trimmed version of GO categories to their human gene elements [@orgHsEg], [@goconsortium2023], [@sample_ects_data].

```{r sample_ects}
data("sample_ects")
sample_ects
```

Object `sample_ects` was trimmed to pass all the suitability tests.
```{r sample_ects_suitability}
system.time(suitability <- summarize_suitability_bipartite(sample_ects))
suitability
```

It was also built to have the same universe of human genes as `sample_ecp` (`sample_ecp <- ECProb(sample_ecg)`). Now, `sample_ecg` and `sample_ects` do not represent the same network: the first one represent interactions between genes/elements and the second one membership of elements in terms.

Methods of `ECTermScoring` are tailored to a bipartite network. This said, an `ECTermScoring` object has an `@ecprob` slot, thus making it possible to also utilize more general `ECProb` methods.

### Test for connectedness of element sets to terms

#### Basic method

The first question to answer is: "Which terms are well related to a given set $E$ of elements?". The task is ranking terms with respect to each other. 

The `ECTermScoring` method to answer this question is `terms_ecset_statistics (obj, set, lambda_method)`

- `obj` is an ECTErmScoring object
- `set` is a character vector that will be reduced to its intersection with `obj@elements`
- `lambda_method` is the type of lambda calculation to be utilized
  - "fast": the default method, with the lowest time complexity
  - "optimized": a hybrid method having intermediate time complexity
  - "accurate": a slow method, used mostly for package development

Here is a first example expected to yield strong connectedness, by using all elements of a term as argument `set`.

```{r ecset_1}
data("sample_ects")
term <- "GO:0005787"
set_1 <- sample_ects@ecprob@adj[[term]]
set_1
dt_1 <- as.data.table(terms_ecset_statistics(sample_ects, set_1))
setorder(dt_1, -log2_Anscombe_ratio)
num_cols <- names(dt_1)[sapply(dt_1, is.numeric)]
dt_1[, (num_cols) := lapply(.SD, signif, digits = 2), .SDcols = num_cols]
dt_1
```

Edge-count statistics are calculated here in the same way they are for between connectedness, with now one of the two sets being a single term. Call $E$ the element set, $k_e$ the degree of element $e$, $k_T$ the degree of term $T$, $M$ the graph size, $X$ the random variable 'edge count between $E$ and $T$ in the RGGED', $m$ its maximum value, and $p_{e,T}$ the RGGED probability of a connection between $e$ and $T$. We have

$$p_{e,T} = \min\left(1, \frac{k_ek_T}{2M}\right)$$
$$m = \left| E \right| \left| T\right| = \left| E \right|$$
$$\lambda = \sum_{e \in E} p_{e,T}$$
$$p_{E,T} = \sum_{u=x}^m \Pr\left(X = u\right)$$
$$\Pr\left(X = u\right) = \frac{e^{-\lambda}}{\alpha}\frac{\lambda^u}{u!}$$
$$\alpha = e^{-\lambda}\sum_{v=0}^{m} \frac{\lambda^v}{v!}$$
$$lAr = \frac{1}{2} \log_2 \frac{x +3/8}{\lambda + 3/8}$$
Again, using $lAr$ as effect size is important because $X$ can have a truncated Poisson distribution of parameter $\lambda$ small enough to yield an asymmetric distribution.

With `lambda_method = "fast"` calculations are simplified as follows

$$p_{ij} = \frac{k_ek_T}{2M} \quad \text{and} \quad \lambda = \frac{k_T}{2M} \sum_{e\in E} k_e$$

Connectedness to terms of `set_1` is strong as shown by large values of `observed_edge_count` and values of `log2_Anscombe_ratio` greater than 1. As expected, the term used to define `set_1` is at the top of the results. 

It is important to stress that results based on an underlying RGGED are not equivalent to those obtained with a Fisher-exact test: the edge-count approach uniquely takes into account the number of terms each element is connected to.

Next are the results obtained with a random set of elements `set_2` having the same size as `set_1`.

```{r ecest_2}
set.seed(2)
set_2 <- sample(sample_ects@elements, length(set_1))
dt_2 <- as.data.table(terms_ecset_statistics(sample_ects, set_2))
setorder(dt_2, -log2_Anscombe_ratio)
num_cols <- names(dt_2)[sapply(dt_2, is.numeric)]
dt_2[, (num_cols) := lapply(.SD, signif, digits = 2), .SDcols = num_cols]
dt_2
```

Results with `set_2` do not indicate strong connectedness to terms.

- `observed_edge_count` values are all 1. Value 1 is the expected minimum, not 0, because random elements are represented in the bipartite graph, i.e. they map to at least one term.
- `log2_Anscombe_ratio` values are positive, but between 0.93 and 0.94, that is, small compared to the best ones seen with `set_1`. Value 0.937 is actually that expected for an observed edge count of 1 and $\lambda$ very small compared to 3/8: $lAr \simeq\log_2\sqrt{1+3/8} / \sqrt{3/8} \simeq 0.937$.
- `p_value` values are larger than those obtained with `set_1`.

Now, it is important to clarify the meaning of `p_value` as returned by method `terms_ecset_statistics`. It means here the probability of observing at least `observed_edge_count` between the selected set and the terms of the bipartite network **under the null model of RGGED** (similar to edges being rewired). `p_value` is a useful measure to compare matched terms, though `log2_Anscombe_ratio` is a better one. Since `sample_ects` contains no isolated elements (every element maps to at least one term), any random set will produce an edge count of at least one and a p-value less than 1. Namely, the distribution of p-values for random sets of elements drawn from `sample_ects` is not uniform.

This distribution can be estimated with a vectorized version of `terms_ecset_statistics`.

#### Vectorized method

Method `terms_ecset_statistics_vectorized` is similar to `terms_ecset_statistics` but it now accepts multiple sets as argument:

- `input_sets`: a set-element data table with columns "set_id" and "element"

Below is the application to 10,000 random sets of the same size as `set_1`.

```{r vectorized_ecset_stats}
data("sample_ects")
n_sets <- 10000
set_size <- length(set_1)

random_elements <- unlist(replicate(n_sets, sample(sample_ects@elements, set_size), simplify = FALSE))

input_sets <- data.table(
  set_id = rep(paste0("S", 1:n_sets), each = set_size),
  element = random_elements
)
system.time(
  result <- terms_ecset_statistics_vectorized(sample_ects, input_sets)
)
```

The returned value `result` is a named list, where each name is a set_id and the value is the corresponding data table of statistics, such as in method `terms_ecset_statistics`, with columns `observed_edge_count`, `log2_Anscombe_ratio` and `p_value`.

```{r term_ecset_stat_control_1}
random_stats_dt <- rbindlist(result, idcol = "set_id")
summary(random_stats_dt$observed_edge_count)
summary(random_stats_dt$log2_Anscombe_ratio)
```

The distributions of `log2_Anscombe_ratio` over 10,000 random sets is indeed centered around 0.934, a smaller value than 0.937 due to $\lambda > 0$ and asymmetry of the distribution. Likewise, the distribution of `p_value` (RGGED p-value) is definitely not symmetric and uniform:

```{r term_ecset_stat_control_2}
summary(random_stats_dt$p_value)
```
```{r term_ecset_stat_control_plot, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=6, fig.align="center", out.width="50%"}

ggplot(random_stats_dt, aes(x = p_value)) +
  geom_density(fill = "#047857", alpha = 0.7, color = NA) +
  labs(title = "Distribution of RGGED P-values for Random Sets",
       x = "p_value (RGGED)",
       y = "Density") +
  theme_minimal(base_size = 22) +
  coord_cartesian(xlim = c(0, 0.014)) 
```

This is of course because only elements connected to at least one term are selected in the random sets. Because the control distribution of `p_value` is not uniform in [0,1], one cannot apply a standard False Discovery Rate (FDR) corrections like Benjamini-Hochberg [@benjamini1995]. Instead, an empirical approach based on simulations is required.

#### FDR correction for connectedness of an element set to terms

The distribution of RGGED p-values for connectedness of random sets to terms is not uniform. This implies that standard False Discovery Rate (FDR) corrections (like Benjamini-Hochberg) are not applicable.To provide a rigorous measure of significance, an Empirical FDR approach is used: compare the scores of the real set against a null distribution generated by drawing random element sets of the same size.

Function `terms_ecset_statistics_fdr` of `ECTermScoring` is a wrapper performing the entire analysis:

- Calculates raw statistics for the input set, as in `terms_ecset_statistics`
- Generates `n_permutations` random sets of the same size, calculates Normalized Enrichment Score (NES), and estimates FDR q-values (function `calculate_empirical_fdr`).

Below is the application to `set_1` (elements of term GO\:0005787).

```{r ecset_fdr_set_1}
# 1000 permutations
system.time(
  fdr_results <- terms_ecset_statistics_fdr(
    sample_ects, 
    element_set = set_1, 
    n_permutations = 1000, 
    seed = 1
  )
)
# Add term names
data("sample_term_lookup")
fdr_results$term_name <- sample_term_lookup[fdr_results$term_id]
# numerical results formatting
num_cols <- names(fdr_results)[sapply(fdr_results, is.numeric)]
fdr_results[, (num_cols) := lapply(.SD, signif, digits = 4), .SDcols = num_cols]

fdr_results
```

The result is a data table of terms ranked by descending `log2_Anscombe_ratio`. The `nes` column indicates the strength of the enrichment relative to the null model, and `fdr_q_value` provides the estimated probability that the finding is a false discovery.

- GO\:0005787 (Signal Peptidase Complex) is the term used to generate set_1. It has the strongest signal (lowest `p_value`, highest 'log2_Anscombe_ratio` and smallest `fdr_q_value`), reflecting the perfect match.

- GO\:0006465 (Signal Peptide Processing) is also a top hit. Its higher `lambda` indicates it is a larger term containing all the elements of the first term. Note that its `nes` score is slightly higher compared to `set_1`. This is because larger terms tend to have lower random `log2_Anscombe_ratio`, creating a smaller denominator for the NES calculation.

Unlike element sets representing actual terms, random element sets will not yield low fdr q-values.

```{r ecset_fdr_set_2}
set.seed(3)
set_3 <- sample(sample_ects@elements, length(set_1))
fdr_results <- terms_ecset_statistics_fdr(
    sample_ects, 
    element_set = set_3, 
    n_permutations = 1000, 
    seed = 1
  )
num_cols <- names(fdr_results)[sapply(fdr_results, is.numeric)]
fdr_results[, (num_cols) := lapply(.SD, signif, digits = 4), .SDcols = num_cols]
setorder(fdr_results, fdr_q_value)
fdr_results[1:5]
```

### Vertex Set Enrichment Analysis (VSEA)


Another analysis type in package `EdgeCount` is a test for connectedness between a ranked list of elements and the terms in the bipartite graph. The question asked is: "Are there terms that connect to the top-ranked elements in a significant way - that is, more than expected by chance?". 

The concept of "chance" here corresponds to a null model where the element ranks are uniformly random.

The full process and its associated random experiment are described below:

- **Conditioning**: A list of element ranks is provided as input. The universe of the analysis (terms and elements) is reduced to the intersection of the graph universe and the input elements to ensure valid statistical conditioning.
- **Enrichment Profile** (Running Score): A running enrichment score is calculated by traversing the ranked list from top to bottom. For every rank $i$ (from 1 to $N_{elements}$):
  - Take the set of the top $i$ elements.
  - Calculate the expected ($\lambda_i$) and observed ($x_i$) edge counts  between this top set and the each term. 
  - Compute the log Anscombe ratio at this rank:
    - $$lAR_i = \frac{1}{2}\log_2 \frac{x_i+3/8}{\lambda_i+3/8}$$
- **Degree Penalization**: A unique property of this profile—unlike standard GSEA—is that it accounts for element degrees. If the top ranks contain "hubs" (elements with high degrees), the expected count $\lambda_i$ increases significantly. This acts as a penalty on the effect size ($lAR$), preventing the identification of enrichment driven solely by promiscuous elements.
- **Null Model**: A control distribution of scores is built by repeating the above process for $N_{perms}$ permutations, where the element ranks, in the reduced universe of elements, are shuffled uniformly at random.
- **Significance**: For each term, estimates of the max (positive enrichment) are computed. FDR q-values are then estimated by comparing the observed NES against the simulated control distribution.

Sorting the results by ascending FDR q-values enables the user to detect terms significantly connected to top ranks (without defining a rank threshold) and below a desired false discovery rate.

Here is a first trivial positive control for method `run_vsea`. Elements of a term of size 5 are given the top ranks.

```{r vsea_pos1}
term_id <- "GO:0000347"
term_elements <- sample_ects@ecprob@adj[[term_id]]
element_ranks <- c(sample(term_elements), sample(setdiff(sample_ects@elements, term_elements)))
system.time(
  vsea_results <- run_vsea(sample_ects, element_ranks = element_ranks, n_permutations = 1000, seed = 1)
)
max_summary <- vsea_results$max_score_summary
max_summary$term_name <- sample_term_lookup[max_summary$term_id]
setorder(max_summary, fdr_q_value)
num_cols <- names(max_summary)[sapply(max_summary, is.numeric)]
max_summary[, (num_cols) := lapply(.SD, signif, digits = 3), .SDcols = num_cols]
head(max_summary, 5)
```

Using `run_vsea` correctly identifies the selected term `GO:0000347` based on its element ranks, as well as two related terms which have common elements.


Next is the obvious negative control of using random element ranks.

```{r vsea_neg}
element_ranks <- sample(sample_ects@elements)
vsea_results <- run_vsea(sample_ects, element_ranks, 1000, seed = 1)
max_summary <- vsea_results$max_score_summary
max_summary$term_name <- sample_term_lookup[max_summary$term_id]
setorder(max_summary, fdr_q_value)
num_cols <- names(max_summary)[sapply(max_summary, is.numeric)]
max_summary[, (num_cols) := lapply(.SD, signif, digits = 3), .SDcols = num_cols]
head(max_summary, 5)
```

This time, no term obtains a small fdr q-value.

Last is a more complex positive control based on both `sample_ects` (term-element network) and `sample_ecg` (element interaction network). The element of highest degree in the interaction network is identified and its set of neighbors is retrieved. These neighbors, not containing the hub itself to avoid trivial enrichment, are then assigned the top ranks.

```{r vsea}
# finds the sample_ecg hub 
data("sample_ecg") 
data("sample_gene_symbol_lookup")
all_degrees <- unlist(sample_ecg@degrees)
gene_id <- names(all_degrees)[which.max(all_degrees)]
gene_symbol <- sample_gene_symbol_lookup[gene_id]
paste0("Hub : ", gene_symbol, " (Entrez: ", gene_id, ")")
# neighborhood ranking
all_elements <- sample_ects@elements
neighbors <- get_neighbors(sample_ecg, gene_id)
valid_neighbors <- intersect(neighbors, all_elements)
ordered_ids <- c(        
  sample(valid_neighbors),                      # neighbors, without hub
  sample(setdiff(all_elements, valid_neighbors)) # the rest
)
element_ranks <- setNames(seq_along(ordered_ids), ordered_ids)
# run VSEA, 1000 permutations
system.time(
  vsea_results <- run_vsea(sample_ects, element_ranks = element_ranks, n_permutations = 1000, seed = 1)
)
# results
max_summary <- vsea_results$max_score_summary
max_summary$term_name <- sample_term_lookup[max_summary$term_id]
setorder(max_summary, fdr_q_value)
num_cols <- names(max_summary)[sapply(max_summary, is.numeric)]
max_summary[, (num_cols) := lapply(.SD, signif, digits = 3), .SDcols = num_cols]
head(max_summary, 8)
```

Using `run_vsea` yields significantly-scoring terms that are related to known functions of the hub 'Vasoactive Intestinal Peptide' (VIP).

# References

