---
title: "EdgeCount Package Tutorial"  
author: "Jo&euml;l R. Pradines"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{EdgeCount Package Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{css, echo=FALSE}
:target {
  background-color: #ffeb3b; /* A nice yellow color */
  transition: background-color 0.8s ease;
}
```

```{r setup}
# library(EdgeCount)
```

```{css, echo=FALSE}
body {
  font-size: 110%;
  line-height: 1.6;
  max-width: 940px;
  margin: 0 auto;
  padding: 20px;
}
```

# Introduction
The EdgeCount package provides a statistical framework for enrichment analysis based on graph connectedness. It is designed for large, sparse graphs, such as protein interaction networks, social networks, or term-element membership graphs (bipartite graphs).

The package's methodology combines the Random Graph with Given Expected Degrees (RGGED) model [[1]](#ref1) with analytical expressions for connectedness effect size and p-value [[2]](#ref2). 

It also introduces Vertex Set Enrichment Analysis (VSEA), a permutation-based method similar to GSEA [[3]](#ref3). VSEA uniquely accounts for the degree of each element (e.g., the number of terms an element belongs to).

The primary functions allow users to:

- Test for internal connectedness within one or more vertex sets.
- Test for connectedness between pairs of vertex sets.
- Score all terms against a single element set or a collection of element sets.
- Perform Vertex Set Enrichment Analysis (VSEA) on a ranked list of elements.
- Summarize VSEA results with Normalized Enrichment Scores (NES) and FDR q-values.

<br>

# Installation

The EdgeCount package is currently available on GitHub. You can install it using the devtools package. If you do not have devtools installed, you can install it from CRAN.

```{r installation, eval=FALSE}
# install.packages("devtools")
devtools::install_github("JoelRMath/EdgeCount")
```

<br>

# References

<a id="ref1">(1.)</a> Chung & Lu (2002). The average distances in random graphs with given expected degrees. *PNAS*, 99 (25) 15879-15882. 
<a href="https://www.pnas.org/doi/10.1073/pnas.252631999" target="_blank">[link]</a>

<a id="ref2">(2.)</a> Pradines, J.R., et al. (2005). Analyzing Protein Lists with Large Networks: Edge-Count Probabilities in Random Graphs with Given Expected Degrees. *Journal of Computational Biology*, 12(2), 113-28.<a href="https://www.liebertpub.com/doi/10.1089/cmb.2005.12.113" target="_blank">[link]</a>


<a id="ref3">(3.)</a> Subramanian, A., et al. (2005). Gene set enrichment analysis: a knowledge-based approach for interpreting genome-wide expression profiles. *PNAS*, 102(43), 15545-50. <a href="https://www.pnas.org/doi/10.1073/pnas.0506580102" target="_blank">[link]</a>


```{r stop_here, echo=FALSE, results='hide'}
knitr::knit_exit()
```


```{r show_categorized_table, echo=FALSE}
# This code assumes the categorized file is now included in your
# package's `inst/extdata` directory for use in the vignette.

# For the vignette, we would read the file like this:
# categorized_pairs <- data.table::fread(
#   system.file("extdata", "term_pairs_categorized.tsv", package = "EdgeCount")
# )

# For now, let's just read it from the data-raw folder for this example
categorized_pairs <- data.table::fread("../data-raw/term_pairs_categorized.tsv")


# --- Use kableExtra to create the color-coded table ---

# Define a color palette for your categories
# You can customize these colors
palette <- c(
  "Complex/Subunit" = "#F8766D",
  "Metabolism" = "#7CAE00",
  "Signaling" = "#00BFC4",
  "Development" = "#C77CFF",
  "Other" = "#A9A9A9"
)

# Display the top 20 pairs
knitr::kable(
  head(categorized_pairs, 20),
  caption = "Top 20 Most Interconnected Term Pairs"
) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::row_spec(
    which(categorized_pairs$category[1:20] == "Complex/Subunit"), 
    background = palette["Complex/Subunit"]
  ) %>%
  kableExtra::row_spec(
    which(categorized_pairs$category[1:20] == "Metabolism"), 
    background = palette["Metabolism"]
  ) %>%
  kableExtra::row_spec(
    which(categorized_pairs$category[1:20] == "Signaling"), 
    background = palette["Signaling"]
  ) %>%
    kableExtra::row_spec(
    which(categorized_pairs$category[1:20] == "Development"), 
    background = palette["Development"]
  ) %>%
  kableExtra::row_spec(
    which(categorized_pairs$category[1:20] == "Other"), 
    background = palette["Other"]
  ) %>%
  kableExtra::scroll_box(width = "100%")
```
